\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This thesis presents the design, implementation, and evaluation of an AI tutor for the Tunisian Baccalaureate mathematics examination (Section Math\'{e}matiques), with a particular focus on Retrieval-Augmented Generation (RAG). The system combines a curated corpus of digitized past examinations, exercise series, and course material with a two-stage retrieval pipeline and a large language model to generate answers intended to remain faithful to the curriculum and to the style of official Bac corrections.

Three approaches were implemented and compared: (1)~a RAG system that grounds answers in retrieved documents, (2)~a prompt-only baseline that relies on the model's internal knowledge without retrieval, and (3)~a hybrid system that adapts its behavior based on retrieval confidence. All three systems use the same underlying generative model, the same persona (``Monsieur Tounsi''), and the same answer format, so the main experimental difference is how domain knowledge is supplied at inference time.

A blind evaluation was conducted by a Tunisian Bac mathematics teacher on 15 questions. The anonymized outputs were graded on four criteria: mathematical correctness, reasoning clarity, pedagogical quality, and adherence to Bac style. The results showed that the prompt-only baseline (mean 4.32/5) outperformed both the RAG system (3.22/5) and the hybrid system (3.25/5). All three systems also correctly refused all four out-of-scope guardrail questions. In addition, a small ground-truth evaluation of the digitization pipeline reported an average Character Error Rate of 5.4\% and a \LaTeX{} math expression accuracy of 97.1\%. Retrieval analysis helps explain these findings. Although retrieval distances were consistently low, chapter-level relevance was correct in only 9 of 15 analyzed queries, showing that low embedding distance did not always correspond to topically appropriate context. This suggests that, in the current setup, retrieval sometimes introduced misaligned material rather than improving the answer. By contrast, the prompt-only system benefited from the strong parametric knowledge of the underlying model. More broadly, the results indicate that retrieval is not automatically beneficial in small, specialized educational corpora, even when the retrieval metrics appear strong.

The thesis contributes a digitized corpus of Tunisian Bac mathematics materials, a domain-specific retrieval architecture with paired exercise--correction retrieval, and an empirical comparison of prompt-only, RAG, and hybrid approaches in a real educational setting. The findings show that careful prompt design can outperform retrieval-based grounding when the base model is strong, retrieval relevance remains imperfect, and the corpus is limited.
