\chapter{Background}
\label{sec:background}
This section introduces the concepts needed to understand the systems developed
in this thesis. We will cover large language models, embedding representations, retrieval-augmented generation,Prompt Engineering,Hybrid AI Architectures, vector databases, and the use of AI in mathematics education .

\subsection{Large Language Models}
\label{subsec:llm-background}

\subsubsection{Transformer Architecture}

Modern large language models are largely based on the Transformer architecture
introduced by Vaswani et al.~\cite{vaswani2017attention}. Unlike earlier
recurrent neural networks, which process text sequentially, Transformers can
process all tokens in a sequence in parallel. This is made possible by the
self-attention mechanism, which allows the model to capture dependencies
between tokens even when they are far apart in the input.

The original Transformer consists of encoder and decoder blocks built from
multi-head self-attention layers and feed-forward networks. At the core of the
architecture is the attention mechanism, which computes how strongly each token
should relate to the others in the sequence. It is commonly written as

\begin{equation}
\text{Attention}(Q, K, V) = \text{softmax}\left(\frac{QK^T}{\sqrt{d_k}}\right) V
\label{eq:attention}
\end{equation}

where $Q$, $K$, and $V$ denote the query, key, and value matrices, and $d_k$
is the dimension of the key vectors. In multi-head attention, this computation
is performed several times in parallel with different learned projections,
allowing the model to capture different types of relationships within the
same sequence.

\subsubsection{From BERT to Generative Models}

Transformer-based language models are often grouped into encoder-only and
decoder-only architectures, depending on how they are used.

Encoder-only models such as BERT~\cite{devlin2019bert} are  trained
with masked language modeling and are mainly used for language understanding
tasks, including classification, named entity recognition, and the computation
of text embeddings. In this thesis, this distinction is important because
embedding-based retrieval is a central component of the RAG system.

Decoder-only models such as GPT-3~\cite{brown2020gpt3},
LLaMA~\cite{touvron2023llama}, and Gemini~\cite{team2023gemini} are trained
autoregressively to predict the next token in a sequence and are therefore
 well suited to text generation. As such models increased in
scale, they became capable of handling a big range of tasks from
natural-language instructions and in-context examples, often without
task-specific fine-tuning~\cite{brown2020gpt3}.

At the same time, these models can produce fluent but incorrect content ("hallucination"). In an
educational situation, this can be  problematic because students may treat
a convincing explanation as reliable even when the underlying reasoning is
wrong.

\subsubsection{Gemini}

Gemini is a family of multimodal language models developed by
Google~\cite{team2023gemini}. In this thesis, Gemini models are used for two
separate tasks.

First, Gemini is used as part of the digitization pipeline to transcribe
scanned mathematical documents into  \LaTeX. This is relevant for the
present corpus, which contains both printed and handwritten mathematical
material.

Second, Gemini serves as the generation model in all three system variants:
the prompt-only system, the RAG system, and the hybrid system. It receives
either the user question alone or the question  with retrieved data
and produces answers intended to follow the style ( the way of the redaction )  and level of Tunisian Baccalaureate mathematics.

Using the same model family for both tasks simplified the implementation and
fit well with the Vertex AI infrastructure used in the project. This was
primarily a practical design choice rather than a claim being evaluated in the
thesis. 

\subsection{Text Embeddings and Semantic Similarity}
\label{subsec:embeddings-background}

\subsubsection{Dense Vector Representations}

Text embeddings represent variable-length text as fixed-dimensional vectors in
a continuous space. In such a space, texts with similar meaning are expected to
lie closer to each other than unrelated texts. Early approaches relied on
word-level representations such as Word2Vec or GloVe, whereas more recent
methods use Transformer-based encoders to produce embeddings for longer text
segments.

Sentence-BERT~\cite{reimers2019sentence_bert} showed that BERT can be adapted
to produce sentence-level embeddings for which cosine similarity is useful as a
proxy for semantic relatedness. This made neural embedding methods much more
practical for retrieval settings, since document embeddings can be computed
offline and query embeddings at runtime.

\subsubsection{BGE-M3: Multilingual Multi-Granularity Embeddings}

For this thesis, the embedding model selected for the retrieval pipeline was
BGE-M3~\cite{chen2024bge_m3}. Several of its properties were relevant to the
present use case.

First, BGE-M3 is multilingual and supports a large number of languages,
including French and Arabic. This is relevant for Tunisian Baccalaureate
materials, which are primarily written in French but may also contain Arabic
terms or annotations.Also ,  the model was designed to support several retrieval paradigms,
including dense retrieval, sparse retrieval, and ColBERT-style late
interaction~\cite{chen2024bge_m3}. In the present system, however, only its
dense embedding capability is used.Moreover,  BGE-M3 supports relatively long input sequences, which is helpful for mathematical documents that may contain extended exercise statements and worked solutions. In practice, this reduced the risk of truncating relevant material
during indexing.

The choice of BGE-M3 was primarily "pragmatic". Compared with API-based
embedding services, it offered an open-weight model that could be run locally
within the project environment. This simplified experimentation, avoided
repeated external API calls during indexing, and gave greater control over the
embedding pipeline.

\subsubsection{Challenges of Mathematical Embeddings}

Embedding mathematical content introduces difficulties that are less prominent
in general-purpose text retrieval. Mathematical expressions often convey
meaning through structure, for example through fractions, superscripts,
subscripts, or spatial layout, whereas standard embedding models operate on a
linear text representation. As a result, semantically important aspects of a
formula may be harder to preserve when the content is serialized as plain text.

Prior work has explored embedding methods designed more specifically for
mathematical expressions and formula retrieval~\cite{mansouri2019tangent,pfahler2020math_embeddings}.
However, such approaches are not always straightforward to integrate into a
corpus like the one used in this thesis, which combines natural-language
exercise statements, worked solutions, and mathematical notation in the same
documents.

In the present system, this issue is addressed pragmatically by digitizing the
documents into \LaTeX\ before indexing. \LaTeX\ provides a standardized linear
representation of mathematical expressions, for example
\texttt{\textbackslash frac\{a\}\{b\}} for $\frac{a}{b}$, which can then be processed by
a general-purpose embedding model. This does not fully resolve the broader
problem of mathematical semantic representation, but it provides a practical
format for retrieval in a mixed text-and-formula corpus. In addition, the
retrieval pipeline does not rely on embeddings alone, since metadata such as
chapter, document type, and solution status are also used to guide document
selection. 

\subsection{Retrieval-Augmented Generation}
\label{subsec:rag-background}

\subsubsection{Motivation and Core Idea}

Retrieval-Augmented Generation (RAG) was introduced by Lewis et
al.~\cite{lewis2020rag} as a way to combine language generation with
retrieval from an external knowledge source. Instead of relying only on the
information stored in model parameters, a RAG system can retrieve relevant
documents at very short time and use them as additional context for better answer
generation.

The basic RAG pipeline consists of three stages:
\begin{enumerate}
    \item \textbf{Indexing}: Documents are split into chunks, embedded as
    vectors, and stored in a retrieval system such as a vector database.
    \item \textbf{Retrieval}: Given a user query, the system retrieves the
    most relevant chunks by comparing the query representation with stored
    document representations.
    \item \textbf{Generation}: The retrieved material is provided to a
    language model together with the user query, and the model generates an
    answer based on this context.
\end{enumerate}

In educational situation, this approach is very attractive because it
allows the model to rely on a controlled corpus rather than only on its
parametric knowledge. For the present thesis, this is important because the
goal is not only to answer mathematics questions correctly, but to generate
solutions that follow the style and methods expected in Tunisian
Baccalaureate corrections.

\subsubsection{Retrieval Methods}

Two common retrieval paradigms are sparse retrieval and dense retrieval.

\begin{itemize}
    \item \textbf{Sparse retrieval} methods such as
    BM25~\cite{robertson2009bm25} rely on lexical overlap and
    term-frequency-based scoring. They are efficient and interpretable, but
    they are limited when relevant documents use different wording from the
    query.

    \item \textbf{Dense retrieval} methods such as
    DPR~\cite{karpukhin2020dpr} represent queries and documents as dense
    vectors and retrieve them by nearest-neighbor search in embedding space.
    These methods are suited to capturing semantic similarity, but
    their performance depends on the quality of the embedding model and the
    fit between the retrieval corpus and the query distribution.
\end{itemize}

Many modern retrieval systems combine lexical and semantic signals. In this
thesis, the retrieval pipeline is primarily dense, using BGE-M3 embeddings.
This is complemented by metadata-based filtering and selection, for example by
document type or chapter, in order to reduce irrelevant retrieval results and
better match the structure of the mathematical corpus.

\subsubsection{Chunking Strategies}

The way a document is divided into chunks has a direct effect on retrieval.
When chunks are too short, relevant context may be split across several units.
When they are too long, retrieval can become less precise because useful
information is embedded together with too much unrelated material.

This is why chunking should not be seen as a purely technical preprocessing
step. Recent work has shown that retrieval quality often depends on how well
chunk boundaries reflect the internal structure of the source documents. Zhao
et al.~\cite{zhao2024financial_chunking}, for instance, report better
retrieval results when chunking follows natural document boundaries rather than
arbitrary character limits. Chen et al.~\cite{chen2024mog} likewise argue that
different chunk sizes can be useful for different purposes, since smaller
chunks may preserve local details more accurately, while larger ones retain
broader context.

In this thesis, chunking is adapted to the type of document being indexed.

Worked corrections are split into smaller chunks so that individual exercises
and solution steps remain as self-contained as possible. Course material, by
contrast, is chunked more coarsely. Here, larger chunks are preferable because
definitions, theorems, and their surrounding explanations often need to stay
together in order to remain interpretable during retrieval. This decision was
mainly implementation-driven, but it is consistent with the broader idea that
chunk size should reflect document structure rather than be fixed uniformly
across the entire corpus.

\subsubsection{Advanced RAG Architectures}

Since the original RAG model, a wide range of retrieval-augmented
architectures has been proposed~\cite{gao2024rag_survey}. These systems differ
mainly in where retrieval is introduced, how tightly it is integrated with the
language model, and how much control the model has over the retrieval process.

Some approaches integrate retrieval deeply into training. RETRO, for
example, incorporates large-scale retrieval during pre-training and shows that
external memory can improve generation beyond what is achieved by parameter
scaling alone~\cite{borgeaud2022retro}. Atlas follows a different direction by
jointly training the retriever and the generator, which allows retrieval to be
optimized more directly for downstream generation tasks~\cite{izacard2022atlas}.

Other work focuses less on changing the base architecture and more on making
retrieval itself more selective. In such systems, the model may decide whether
retrieval is needed, assess the usefulness of the retrieved material, or adjust
its generation strategy accordingly.

The hybrid system used in this thesis is much simpler than these research
architectures. It does not introduce a new retriever, nor does it jointly
train retrieval and generation. Instead, it uses a practical routing mechanism
at inference time. Retrieval is attempted first, and the system then decides
whether to rely mainly on retrieved documents, combine retrieval with
parametric model knowledge, or fall back to prompt-only generation.

This design was motivated by a concrete issue in the present corpus:
retrieval is helpful when the retrieved material is genuinely relevant, but
forcing weak matches into the prompt can degrade answer quality rather than
improve it.

\subsection{Prompt Engineering}
\label{subsec:prompt-engineering}

Prompt engineering refers to the design of the input given to a large language
model in order to influence the form, style, and quality of the generated
output. Unlike fine-tuning, prompt engineering does not modify the model
parameters. It operates entirely at inference time and is therefore much
lighter to iterate on in practice~\cite{white2023prompt_patterns}.

\subsubsection{Fundamental Prompting Strategies}

A useful distinction can be made between zero-shot, few-shot, and
reasoning-oriented prompting strategies.

In zero-shot prompting, the model receives only the task description.
Few-shot prompting extends this setup by adding a small number of input-output
examples before the actual query. Brown et al.~\cite{brown2020gpt3} showed
that such examples can substantially improve performance without any gradient
updates, simply by using the prompt itself as a form of task specification.

For reasoning tasks, prompting strategies often go a step further by asking
the model to produce intermediate steps rather than only a final answer.
Kojima et al.~\cite{kojima2022zero_shot_cot} found that even a simple cue such
as "Let´s think step by step" can improve performance on multi-step tasks.
Wei et al.~\cite{wei2022chain_of_thought} formalized this idea in
chain-of-thought prompting, where the model is encouraged to generate explicit
reasoning traces. This is particularly relevant for mathematical problem
solving, where the quality of an answer depends not only on the final result
but also on the correctness and clarity of the intermediate steps.

Instruction-tuned models add another layer to this picture. Ouyang et
al.~\cite{ouyang2022instructgpt} showed that models trained to follow natural
language instructions can respond more reliably to structured user requests.
This matters for the present thesis because all three systems rely, in
different ways, on prompts that specify not only what to answer, but also how
the answer should be written.

\subsubsection{Prompt Engineering for Mathematical Tasks}

Prompting mathematical tasks is more demanding than prompting general text
generation. A useful mathematical answer must usually respect notation,
logical order, and the level of detail expected by the user. In an
educational situation , it is often not enough for the final result to be
correct; the method also has to be understandable and written in an acceptable
style~\cite{imani2023mathprompter}.

For this reason, prompts for mathematical tutoring typically need to do more
than state the question. They may specify the notation format, request
step-by-step reasoning, require explicit justification of transformations, or
constrain the method to a given curriculum. In the case of the Tunisian
Baccalaureate, this includes features such as formal French mathematical
phrasing, explicit logical connectors, and the avoidance of methods that lie
outside the expected syllabus.

Automatic Prompt Engineering (APE) explores whether prompts themselves can be
generated and optimized automatically~\cite{zhou2023ape}. While this idea is
interesting, the present thesis relies on manually designed prompts. The main
reason is that the target style is highly specific: the goal is not generic
mathematical helpfulness, but answers that resemble Tunisian Bac corrections
in structure, tone, and level of explanation.

\subsubsection{Role of Prompt Engineering in RAG Systems}

In a retrieval-augmented system, prompting has an additional function. The
prompt must not only describe the task, but also tell the model how to use the
retrieved material. This includes, for example, instructing the model to rely
on the provided context, to cite or mention the relevant sources, and to state
clearly when the available material is insufficient~\cite{gao2024rag_survey}.

In this thesis, prompt engineering is used in all three system variants, but
its role differs across them. In the prompt-only baseline, the prompt carries
the entire burden of enforcing style and pedagogical structure. In the RAG and
hybrid systems, the prompt also has to organize the interaction between the
user question and the retrieved mathematical documents.

\subsection{Hybrid AI Architectures}
\label{subsec:hybrid-background}

In this thesis, the term hybrid refers to a system that does not rely on a
single response strategy for every query. Instead, it combines retrieval-based
and prompt-based generation and chooses between them depending on the quality
of the retrieved evidence.

\subsubsection{Hybrid Retrieval}

A common use of the term hybrid retrieval refers to the combination of sparse
and dense retrieval methods~\cite{lin2021hybrid_retrieval}. The motivation is
straightforward: sparse retrieval is often strong when exact lexical matching
matters, whereas dense retrieval is better at capturing semantic similarity
across different phrasings. Lin et al.~\cite{lin2021hybrid_retrieval} show
that combining both signals can outperform either one on its own in many
settings.

The hybrid system studied in this thesis is not a classic sparse+dense
retrieval system. Its retrieval component remains primarily dense. However,
document selection is still guided by additional structured signals, in
particular metadata such as chapter, document type, and solution status. In
that limited sense, retrieval is not purely semantic, but constrained by the
structure of the corpus.

\subsubsection{Adaptive and Conditional Generation}

Hybrid ideas also appear at the generation level. Rather than always applying
the same retrieval-and-generation pipeline, some systems try to decide when
retrieval is needed and how strongly the model should rely on it.

Self-RAG is one example of this direction. Asai et al.~\cite{asai2023self_rag}
propose a framework in which the model learns to reflect on whether retrieval
is necessary, whether the retrieved passages are relevant, and whether the
generated answer is supported by the available evidence. Other work has
explored similarly selective designs in which a routing component decides
whether a query should go through retrieval at all~\cite{jiang2023active_rag}.

These systems are more sophisticated than the one implemented in this thesis,
but they motivate the same general intuition: retrieval is useful when it adds
relevant evidence, but it should not be forced indiscriminately for every
query.

\subsubsection{Confidence-Based Routing}

The hybrid engine used in this thesis follows a simple adaptive strategy. It
first performs retrieval and then evaluates the quality of the retrieved
documents using similarity scores. If retrieval appears strong enough, the
system behaves like a standard RAG pipeline. If retrieval is weak but not
entirely irrelevant, it combines retrieved material with prompt-based guidance.
If retrieval is poor, it falls back to prompt-only generation.

This design was motivated by a practical observation during development.
Retrieved documents are not always helpful. In some cases, especially when the
match is weak or belongs to the wrong chapter, forcing those documents into the
prompt can degrade the final answer instead of improving it. The hybrid system
was introduced to avoid that failure mode while still benefiting from
retrieval when the retrieved material is genuinely useful.

This idea is related in spirit to the broader question of when retrieval is
actually necessary~\cite{mallen2023trust_llm}. In the present educational
setting, some standard mathematical knowledge may already be available in the
model's parametric memory, whereas specific Bac-style exercise solutions are
more likely to require retrieval from the curated corpus. The hybrid system is
therefore best understood as a practical routing strategy rather than as a new
retrieval architecture.


\subsection{Vector Databases}
\label{subsec:vector-db-background}

Vector databases are storage systems designed for similarity search over
high-dimensional vector representations. Unlike traditional databases, which
are mainly optimized for exact matches or relational queries, vector databases
support nearest-neighbor search in embedding space and are therefore commonly
used in retrieval pipelines for language models.

A number of systems can be used for this purpose, including
FAISS~\cite{johnson2019faiss}, Pinecone, Weaviate, and
ChromaDB~\cite{chromadb2024}. In this thesis, ChromaDB was selected mainly
for practical reasons.

First, it supports local persistence, which made it possible to keep the full
retrieval pipeline inside the project environment without depending on an
external service. Second, it provides flexible metadata handling, which is
important for the present corpus because retrieval is not based on semantic
similarity alone. Metadata such as chapter, document type, and solution status
also play a role in document selection. Finally, ChromaDB integrates easily
with a Python-based workflow and allows the use of a custom embedding model,
which made it a convenient fit for the implemented system.

\subsection{OCR and Mathematical Document Processing}
\label{subsec:ocr-background}

Converting scanned mathematical documents into machine-readable text is a
well-known challenge in document analysis~\cite{zanibbi2016math_retrieval}.
Standard OCR systems designed for running text often struggle with
mathematical notation, since formulas are not purely linear. Fractions,
superscripts, subscripts, matrices, and other expressions depend on spatial
layout, which makes mathematical OCR substantially harder than plain-text
recognition.

Specialized tools such as Mathpix were developed precisely for this kind of
task~\cite{mathpix2024}. More recently, multimodal models have also shown
promising performance on OCR-like tasks involving mathematical and visually
structured documents~\cite{wang2023ocrbench,mahdavi2019icdar_math}.

In this thesis, digitization is performed with Gemini rather than with a
dedicated mathematical OCR system. This was primarily a practical choice.
The corpus contains not only formulas, but also French explanatory text,
exercise layouts, and in some cases handwritten material or annotations.
Using a multimodal model made it possible to handle these mixed inputs within
a single pipeline. This does not imply that Gemini is universally superior to
specialized math-OCR systems, but it proved suitable for the particular
document collection used in this project.

\subsection{AI in Mathematics Education}
\label{subsec:ai-education}

The use of AI in education has expanded rapidly with the spread of large
language models and related interactive systems~\cite{kasneci2023chatgpt_education}.
In mathematics education, current tools differ not only in technical design,
but also in the kind of support they provide.

General-purpose language models such as ChatGPT, Gemini, or Claude can often
produce mathematically plausible answers, but they are not tied to a specific
curriculum and may use methods or notation that do not match a given school
context. By contrast, mathematics-oriented tools such as Wolfram Alpha or
Photomath are stronger in symbolic computation and procedural solutions, but
they are not designed around the conventions of a particular national exam
system.

This thesis is motivated by that gap in practice. The aim is to build a system
for Tunisian Bac mathematics that does more than answer questions correctly:
it should also follow the expected correction style, remain within curriculum
boundaries, and make use of authentic educational material through retrieval.
