% =============================================================================
%  CHAPTER 6 — CONCLUSION (< 1 page)
% =============================================================================
\section{Conclusion}
\label{sec:conclusion}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Summary}
\label{subsec:summary}

This thesis presented the design, implementation, and evaluation of a
Retrieval-Augmented Generation system for the Tunisian Baccalaureate
mathematics examination (Section Mathématiques). We developed a complete
pipeline—from physical document acquisition and multimodal OCR digitization
through adaptive chunking and vector indexing to interactive, bilingual
answer generation—and compared three system variants: a RAG system grounded
in retrieved curriculum documents, a prompt-only baseline encoding the
syllabus directly into the system prompt, and a hybrid engine that
dynamically routes between the two strategies based on retrieval quality.

A blind evaluation by a Tunisian Baccalaureate mathematics teacher, grading
anonymized outputs across 15 questions on four criteria, produced a result
that ran counter to our initial expectations: the Prompt-Only baseline
(mean 4.32/5) outperformed both the RAG system (3.22/5) and the Hybrid
engine (3.25/5) on every criterion. All three systems passed all four
curriculum guardrail tests, correctly refusing out-of-scope methods. A
separate ground truth evaluation of the digitization pipeline confirmed an
average Character Error Rate of 5.4\% and a \LaTeX{} math expression
accuracy of 97.1\%, indicating that the Gemini-based OCR produces largely
faithful transcriptions but that occasional errors in mathematical notation
can propagate into the retrieved context.

These results suggest that, for a domain where the generative model's
parametric knowledge already covers the mathematical content, and where the
retrieval corpus is modest in scale, careful prompt engineering can be more
effective than retrieval augmentation. The value of RAG appears to depend
not only on the quality of the retrieval pipeline but also on whether the
corpus is large and precise enough to consistently outperform what the
model already knows.

The system addresses a practical gap in educational technology for Tunisian
secondary school students, providing the first AI tutoring tool specifically
designed for the Tunisian Bac mathematics curriculum and its distinctive
correction style. The complete pipeline—including the digitized corpus, the
retrieval engine, and all three system variants—is implemented and
functional, and can serve as a foundation for future work.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Future Work}
\label{subsec:future-work}

Several directions for future work emerge from this thesis:

\begin{enumerate}
    \item \textbf{Corpus expansion}: The most direct way to test whether
    retrieval can surpass prompt engineering is to scale the corpus
    significantly—more years of past examinations, more exercise series, more
    correction variants across all chapters. A larger, denser corpus would
    increase the likelihood of retrieving closely matching corrections, which
    is the scenario where RAG should provide the greatest benefit.

    \item \textbf{Multi-rater evaluation}: Repeating the blind evaluation
    with multiple independent teachers would allow inter-rater reliability
    analysis and strengthen the statistical basis of the comparison. A
    sample of 30--50 graded questions, scored by at least two evaluators,
    would permit meaningful significance testing.

    \item \textbf{Student user study}: The current evaluation measures answer
    quality as judged by a teacher. A complementary study with actual
    Tunisian Baccalaureate students could assess whether the system improves
    learning outcomes, which is the ultimate measure of a tutoring tool's
    value.

    \item \textbf{Embedding fine-tuning}: Fine-tuning the BGE-M3 embedding
    model on Tunisian Bac mathematical text could sharpen retrieval
    precision at the critical boundary between relevant and irrelevant
    documents, potentially improving RAG and Hybrid system performance
    without requiring a larger corpus.

    \item \textbf{Automated evaluation}: Developing automated metrics for
    mathematical correctness and style adherence—potentially using an
    LLM-as-judge approach with a stronger model—would enable larger-scale
    evaluation without the bottleneck of manual expert grading.

    \item \textbf{Interactive tutoring}: The current system produces
    complete answers in a single turn. Adding step-by-step guided problem
    solving, where the system reveals the solution incrementally and
    responds to student attempts, would move the system from a correction
    generator toward an interactive tutor.
\end{enumerate}
