% =============================================================================
%  CHAPTER 5 — DISCUSSION (~10-15% of thesis)
% =============================================================================
\section{Discussion}
\label{sec:discussion}

The blind evaluation produced a clear ranking: the Prompt-Only baseline
outperformed both the RAG system and the Hybrid engine on every criterion
and in every question category. This section examines why these results
emerged, what they imply about the value of retrieval in this domain, and
what limitations constrain the conclusions that can be drawn.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Interpretation of Results}
\label{subsec:interpretation}

\subsubsection{Why the Prompt-Only Baseline Performed Best}

The most striking result is that the system with no retrieval at all
received the highest scores on all four criteria. Three factors likely
contribute to this outcome.

First, the generative model used in all three systems—Gemini
2.5~Flash~\cite{team2023gemini}—is a state-of-the-art large language model
trained on a broad multilingual corpus that almost certainly includes
substantial French-language mathematical content. Standard Tunisian
Baccalaureate topics (complex numbers, integration by parts, sequences and
convergence, probability) are well-represented in the mathematical
literature and in online educational resources. For these standard topics,
the model's parametric knowledge is likely sufficient to produce correct
solutions without external retrieval. This aligns with observations by
Brown~et~al.~\cite{brown2020gpt3} that large language models internalize
considerable factual and procedural knowledge during pretraining.

Second, the Prompt-Only system operates without the constraint of
conditioning on retrieved context. The RAG system's anti-hallucination
instructions explicitly tell the model to use \textit{only} the retrieved
documents. When the retrieved context is incomplete—a correction that
covers only part of the exercise, or a course excerpt that lacks the
specific theorem needed—this constraint forces the model to work within a
limited and potentially misleading context rather than drawing on its full
parametric knowledge. The Prompt-Only system faces no such constraint: it
can freely combine its knowledge of the curriculum with its general
mathematical competence.

Third, the ten-component prompt architecture
(Section~\ref{sec:prompt-only}) provides structured guidance without
constraining the knowledge source. The curriculum inventory tells the model
\textit{what topics} to cover; the reasoning guardrails tell it
\textit{how to structure} its reasoning; the hallucination controls tell it
\textit{when to express uncertainty}. These instructions steer the model
effectively without the overhead and potential noise of retrieval.


\subsubsection{Why RAG and Hybrid Scored Lower}

The RAG system's lower scores suggest that, at the current corpus scale
(903 chunks across 15+ chapters), retrieval introduces more noise than
signal for many queries. The retrieval routing analysis
(Section~\ref{subsec:retrieval-routing-analysis}) provides concrete evidence
for this interpretation: while all 19 queries achieved best L2 distances
below 1.0—well within the strong-match threshold—only 8 of the 19 (42\%)
retrieved documents from the correct mathematical chapter. Low embedding
distance, in other words, did not reliably indicate topical relevance.

This disconnect likely arises from the structure of the corpus. A substantial
proportion of the indexed documents are multi-topic Baccalaureate exam
compilations (``Bac avec corrections Images'') that span several chapters
within a single file. Because these documents contain diverse mathematical
vocabulary, they score low L2 distances against many queries regardless
of the specific topic being asked about. When such a document is retrieved
and presented as authoritative context, the model is instructed to follow it
closely (Case~A anti-hallucination rules), even though the relevant
portion of the document may address a different exercise entirely. The model
must then attempt to reconcile the retrieved context with the actual
question, which can produce a confused or incomplete answer.

The qualitative examples in Section~\ref{subsec:qualitative-examples}
illustrate this concretely. For Question~B02 (probability), the RAG system
achieved its best distance across the entire evaluation ($d = 0.40$), yet
the retrieved corrections addressed simultaneous draws rather than the
sequential draws asked for. For Question~C01 (continuity at a point), the
retrieved material covered continuity on intervals—a related but distinct
concept. In both cases, the model's output was constrained by partially
relevant context rather than enhanced by it.

\subsubsection{Why the Hybrid Engine Did Not Differentiate}
\label{subsec:retrieval-routing-discussion}

The Hybrid engine scored nearly identically to the standalone RAG system
(3.25 vs.\ 3.22). The retrieval analysis reveals a structural reason for
this: all 19 queries were classified as Case~A by both engines, meaning the
Hybrid engine's distinctive routing behavior—Cases~B (partial match) and~C
(no match), which merge retrieved context with curriculum knowledge or fall
back to pure prompt engineering—was never activated during the evaluation.

This outcome was not anticipated in the system design. The thresholds
(Case~A $\leq 1.2$, Case~B $\leq 1.6$) were set based on preliminary
testing, but the evaluation revealed that the BGE-M3 embedding model
produces distances below 1.0 for virtually any query against a corpus of
this size and diversity. The threshold values may have been set too
generously, effectively collapsing the three-case architecture into a
single case.

A more aggressive threshold—for example, lowering the Case~A boundary from
1.2 to 0.6 or 0.7—might have routed some queries to Cases~B or~C,
potentially allowing the Hybrid engine to demonstrate its distinctive
contribution. However, lowering the threshold without also improving the
chapter-match precision of the retrieval would risk routing queries away from
Case~A even when relevant documents \textit{were} retrieved. Threshold
calibration and its interaction with corpus composition remain open questions
for future work.

The Category~D (Derja/mixed language) results are instructive in this regard.
Despite both Derja queries being classified as Case~A (distances of 0.89 and
0.96), the Hybrid engine scored 4.12 compared to the RAG system's 3.50 on
these questions. One explanation is that the Derja queries, while achieving
low distances, may have retrieved documents where the language mismatch
(Derja query against French-language corpus) was partially compensated by the
Hybrid engine's curriculum-aware prompt assembly, even within Case~A.

\subsubsection{The Distance--Relevance Gap}

Perhaps the most important finding from the retrieval analysis is the
disconnect between embedding distance and topical relevance. All queries
fell below the strong-match threshold, yet fewer than half retrieved
documents from the correct chapter. This suggests that L2 distance in the
BGE-M3 embedding space is not a reliable proxy for mathematical topic
alignment at the chapter level—at least not at the current corpus scale
and with the current chunking strategy.

Several factors may contribute to this gap. First, the BGE-M3 model was
not fine-tuned on mathematical or Tunisian educational content; it is a
general-purpose multilingual model. Mathematical queries from different
chapters may share substantial vocabulary (``montrer que,'' ``calculer,''
``en déduire que'') that produces similar embeddings even when the
mathematical content differs. Second, the multi-topic exam compilations in
the corpus act as nearest-neighbor ``attractors'' for many queries because
they contain vocabulary from multiple chapters. Third, the chunking strategy
(1{,}500 characters for corrections) may split exercises in ways that
obscure their chapter-specific content.

This finding has practical implications: a retrieval system for mathematical
tutoring may benefit from metadata-aware filtering (restricting retrieval
to the detected chapter) in addition to, or instead of, purely
distance-based selection. Such an approach would require a reliable chapter
detection step before retrieval, which was not implemented in the current
system.

\subsubsection{The Second-Pass Silence}

The two-stage retrieval strategy was designed to provide theorem-level
context when correction-only retrieval was insufficient. In practice, the
second pass (course material) was never activated, because the first pass
always yielded distances below 1.2. This means the LLM was conditioned
exclusively on past exam corrections for all 19 queries, never on theorem
definitions or course notes.

For coaching-mode questions (Category~C), which ask for conceptual
explanations, this is a notable mismatch: the student asks ``what is the
method?'' and the system retrieves worked solutions to different exercises
rather than the underlying definitions and theorems. The Prompt-Only system,
which encodes curriculum structure directly in its prompt, was free to
provide the conceptual explanation the student needed. This likely contributes
to the large Category~C performance gap (Prompt-Only 4.75 vs.\ RAG 3.33).


\subsubsection{Per-Category Observations}

Category~C (informal student questions) showed the largest gap between
the Prompt-Only system (4.75) and the retrieval-based systems (RAG 3.33,
Hybrid 2.92). The retrieval analysis confirms that this is not a retrieval
failure in the distance sense—all three Category~C queries achieved
distances below 0.92—but rather a content-type mismatch: coaching questions
need pedagogical explanations, and the corpus contains only worked
corrections.

Category~B (novel questions) was the only category where the Hybrid engine
(3.30) outperformed the RAG system (3.00) by a meaningful margin. Despite
all Category~B queries falling into Case~A, the chapter match rate was
higher for this category (3 of 5 matched), which may have provided the
Hybrid engine with slightly more relevant context to work with.

The guardrail results (4/4 for all systems) confirm that the syllabus guard
mechanism works as intended regardless of the retrieval strategy. Since all
three systems share the same forbidden-method list in their system prompts,
this result is expected.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Value of Retrieval for Curriculum-Specific Tasks}
\label{subsec:value-of-retrieval}

These results do not imply that retrieval-augmented generation is without
value for educational mathematics tutoring. They suggest, more narrowly,
that \textit{with the current corpus size, embedding model, and generative
model combination}, retrieval did not provide a net benefit over careful
prompt engineering.

The original RAG framework~\cite{lewis2020rag} was designed for
knowledge-intensive tasks where the answer depends on specific facts that
the language model is unlikely to have memorized—rare entities, recent
events, or domain-specific data. The Tunisian Baccalaureate mathematics
curriculum, while specific in its conventions, covers standard mathematical
content that a large language model has likely encountered during
pretraining. The domain specificity lies less in the \textit{mathematical
content} (which is universal) and more in the \textit{redaction style,
curriculum boundaries, and examination conventions} (which are specific to
Tunisia). This distinction may explain why the Bac Style Adherence
criterion showed the smallest performance gap between the Prompt-Only
system (3.73) and the retrieval-based systems (RAG 2.93, Hybrid 3.07):
style is precisely the dimension where retrieved corrections should add
the most value, and indeed the absolute gap is widest here, suggesting
retrieval may have actively interfered with style rather than helping it.

Gao~et~al.~\cite{gao2024rag_survey} observe in their survey of RAG systems
that retrieval quality depends heavily on corpus coverage and embedding
alignment: a RAG system with poor retrieval can perform worse than the
base model alone, because irrelevant context misleads the generator. Our
results are consistent with this observation. The corpus of 1{,}422 chunks
is small by RAG standards, and the BGE-M3 embedding model, while
multilingual, was not fine-tuned on mathematical or Tunisian educational
content.

A larger corpus with broader coverage—more years of examinations, more
exercise series, more correction variants—might shift the balance in
favor of retrieval by reducing the frequency of partial or irrelevant
matches. Similarly, fine-tuning the embedding model on Tunisian Bac
mathematical text could improve retrieval precision at the critical
boundary between relevant and irrelevant documents. These remain open
questions for future work.

The digitization quality evaluation (Section~\ref{subsec:ocr-evaluation})
provides a complementary perspective. With an average CER of 5.4\% and
\LaTeX{} math accuracy of 97.1\%, the corpus is not dominated by OCR
errors—but it is not error-free either. Roughly one in twenty characters
and one in thirty math expressions contain transcription mistakes. These
errors affect only the retrieval-based systems, since the Prompt-Only
baseline never accesses the corpus. While we cannot isolate the exact
contribution of OCR noise to the RAG system's lower scores, it is a
plausible contributing factor: a single corrupted exponent or dropped sign
in a retrieved correction can mislead the generative model's entire
solution.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Limitations}
\label{subsec:limitations}

Several limitations constrain the generality of these findings.

\subsubsection{Single Evaluator}

The most significant limitation is that the entire evaluation was conducted
by a single teacher. While this evaluator has deep domain expertise in
Tunisian Baccalaureate mathematics, a single rater cannot capture the full
range of pedagogical preferences and grading standards. Different teachers
may weigh mathematical rigor versus pedagogical clarity differently, or may
have varying tolerance for stylistic deviations from the official
correction format. Inter-rater reliability, which would require at least
two independent evaluators, could not be assessed in this study. The
results should therefore be interpreted as one expert's assessment rather
than a consensus judgment.

\subsubsection{Sample Size}

The evaluation covered 15 graded questions across four categories. While
these questions were designed to span the curriculum and test different
query types, 15 questions is a small sample for drawing strong statistical
conclusions. The per-category breakdowns (5, 5, 3, and 2 questions
respectively) are too small for meaningful significance testing. The
observed differences—particularly the Hybrid engine's relative strength
on Category~D (2 questions)—should be treated as suggestive rather than
conclusive.

\subsubsection{Corpus Scale}

The evaluation was conducted with a corpus of 1{,}422 chunks derived from
624 source files. This is a modest corpus by RAG standards. The performance
of the RAG and Hybrid systems is likely sensitive to corpus size: a
substantially larger corpus could improve retrieval quality and potentially
change the relative ranking of the systems. Our results reflect the
specific corpus available at the time of evaluation, not an inherent
limitation of retrieval-augmented approaches.

\subsubsection{Model Dependency}

All three systems use Gemini 2.5~Flash as the generative model. A different
model—particularly a smaller or less capable one—might lack the parametric
mathematical knowledge that allows the Prompt-Only baseline to perform well,
potentially favoring retrieval-augmented approaches. The results are
specific to this model and cannot be assumed to hold for other language
models.

\subsubsection{No Student Evaluation}

This evaluation measured answer quality as judged by an expert teacher, not
learning outcomes for students. A system that produces technically correct
answers may not be the most effective for student learning; conversely, a
system with lower mathematical accuracy but better pedagogical scaffolding
might produce better learning outcomes. Evaluating actual student learning
would require a user study with Tunisian Baccalaureate students, which was
outside the scope of this thesis.

\subsubsection{Digitization Quality}

The corpus was digitized using Gemini as a multimodal OCR engine
(Section~\ref{subsec:digitization}). While we manually verified a sample
of the digitized outputs, OCR errors in mathematical notation—a misread
exponent, a dropped minus sign, a garbled fraction—can propagate into the
vector store and corrupt the retrieved context. Such errors would
disproportionately affect the RAG and Hybrid systems, which rely on
retrieved content, while leaving the Prompt-Only system unaffected.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Comparison with Related Work}
\label{subsec:comparison-related-work}

Direct comparison with existing systems is difficult because, to our
knowledge, no prior work has built a RAG-based tutoring system specifically
for the Tunisian Baccalaureate mathematics curriculum. The closest related
efforts fall into two categories.

General-purpose RAG systems such as those surveyed by
Gao~et~al.~\cite{gao2024rag_survey} typically operate on large-scale
knowledge bases (Wikipedia, scientific literature, enterprise documents)
and target factual question answering rather than mathematical solution
generation. Our system differs in that the retrieval target is not factual
information but \textit{stylistic and methodological exemplars}—the goal is
to find a worked correction whose redaction style the model can mimic, not
to retrieve facts that the model does not know.

AI-in-education systems such as those reviewed by
Kasneci~et~al.~\cite{kasneci2023chatgpt_education} have explored LLM-based
tutoring for various subjects, but typically without the
retrieval-augmented component and without targeting a specific national
examination format. Our work contributes the observation that for a
curriculum with strong stylistic conventions, the choice between retrieval
and prompt engineering is not straightforward: a well-engineered prompt can
outperform retrieval when the generative model already possesses the
relevant mathematical knowledge and the corpus is not large enough to
provide consistently high-quality matches.

The adaptive retrieval approaches of CRAG~\cite{yan2024practical_rag} and
Self-RAG~\cite{asai2023self_rag} are architecturally related to our Hybrid
engine. Our results suggest that even with quality-aware routing, the
benefit of retrieval is limited when the base model is strong and the
corpus is small. This observation complements the CRAG and Self-RAG
findings, which were demonstrated on larger-scale retrieval tasks where
the base model's parametric knowledge is less likely to suffice.
