% =============================================================================
%  CHAPTER 4 — EXPERIMENTS (~20-35% of thesis)
% =============================================================================
\section{Experiments}
\label{sec:experiments}

This section describes the experimental setup, evaluation methodology, and
results. Following the supervisor's guidance, we focus on the quality of
implementation and the scientific approach rather than achieving specific
performance targets.

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Experimental Setup}
\label{subsec:experimental-setup}

\subsubsection{Corpus Statistics}

% TODO: Fill in actual numbers from build_db.py --stats output
\begin{table}[H]
\centering
\caption{Corpus statistics after digitization and indexing.}
\label{tab:corpus-stats}
\begin{tabular}{lr}
\toprule
\textbf{Metric} & \textbf{Value} \\
\midrule
Total raw files scanned            & TODO \\
Successfully digitized files       & TODO \\
Total .tex files in GCS            & 624 \\
Total chunks in ChromaDB           & 903 \\
Chapters covered                   & TODO \\
Years of Bac exams included        & TODO \\
Course material chunks             & TODO \\
Correction/exercise chunks         & TODO \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Hardware and Software Environment}

All experiments were conducted on the following environment:
\begin{itemize}
    \item \textbf{Platform}: Google Cloud Vertex AI Workbench
    \item \textbf{Python}: 3.10
    \item \textbf{Embedding model}: BGE-M3 (\texttt{BAAI/bge-m3}), running
    on CPU with FP32 precision
    \item \textbf{Generation model}: Gemini 2.5 Flash via Vertex AI
    \item \textbf{Vector database}: ChromaDB (local persistent storage)
    \item \textbf{Embedding dimension}: 1024 (BGE-M3 dense vectors)
    \item \textbf{Maximum embedding length}: 8192 tokens
\end{itemize}

\subsubsection{Hyperparameters}

Table~\ref{tab:hyperparameters} lists the key hyperparameters used across
all experiments.

\begin{table}[H]
\centering
\caption{System hyperparameters.}
\label{tab:hyperparameters}
\begin{tabular}{llr}
\toprule
\textbf{Component} & \textbf{Parameter} & \textbf{Value} \\
\midrule
\multirow{2}{*}{Chunking (corrections)} & Chunk size & 1500 chars \\
                                          & Overlap & 200 chars \\
\multirow{2}{*}{Chunking (course)}       & Chunk size & 3000 chars \\
                                          & Overlap & 250 chars \\
\midrule
\multirow{4}{*}{Retrieval}               & $k$ first pass & 10 \\
                                          & $k$ second pass & 8 \\
                                          & Top-$n$ to LLM & 6 \\
                                          & Companion chunks & 3 \\
\midrule
\multirow{2}{*}{Thresholds (L2)}         & Good match & $\leq 1.2$ \\
                                          & Fallback match & $\leq 1.6$ \\
\midrule
\multirow{2}{*}{Generation}              & Max context chars & 14000 \\
                                          & Max chars per doc & 2000 \\
\bottomrule
\end{tabular}
\end{table}


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Evaluation Methodology}
\label{subsec:evaluation-methodology}

Evaluating an AI tutoring system for mathematics is inherently challenging.
Unlike natural language generation tasks where automated metrics like BLEU
or ROUGE can serve as rough proxies, mathematical correctness requires
verifying logical validity, computational accuracy, and adherence to
domain-specific conventions that no automated metric captures
reliably~\cite{kasneci2023chatgpt_education}. We therefore designed a
blind human evaluation protocol centered on expert judgment from a
practicing Tunisian Baccalaureate mathematics teacher.

\subsubsection{Question Bank}

The evaluation used 20 questions distributed across five categories, each
designed to probe a different aspect of system behavior:

\begin{itemize}
    \item \textbf{Category~A} (5 questions): Direct Bac-style exercises
    closely resembling real examination problems. These questions are
    likely to have near-exact matches in the corpus, testing whether the
    RAG system can retrieve and leverage relevant corrections.

    \item \textbf{Category~B} (5 questions): Novel chapter-based exercises
    on the same Bac topics but with different numbers, formulations, or
    problem structures. These test whether the systems can generalize
    beyond memorized examples.

    \item \textbf{Category~C} (3 questions): Informal student-style
    questions written as a real Tunisian student might type them—incomplete
    phrasing, colloquial French, conceptual rather than computational. These
    test pedagogical quality and coaching ability.

    \item \textbf{Category~D} (2 questions): Questions in Tunisian Arabic
    (Derja) or mixed French-Derja, testing language robustness and the
    system's ability to handle code-switching.

    \item \textbf{Category~E} (4 questions): Out-of-scope questions
    requesting methods explicitly outside the Tunisian Bac program
    (L'Hôpital's rule, matrix diagonalization, Taylor series, Lebesgue
    integration). These test the curriculum guardrail mechanism. Unlike
    categories A--D, these are evaluated on a binary pass/fail basis rather
    than a graded rubric.
\end{itemize}

Categories A through D (15 questions) were evaluated using a four-criterion
grading rubric. Category~E (4 questions) was evaluated separately as a
guardrail test.

\subsubsection{Grading Rubric}

Each of the 15 graded questions was scored independently for each system
on four criteria, each on a 0--5 scale:

\begin{enumerate}
    \item \textbf{Mathematical correctness} (0--5): Are the computations and
    reasoning steps mathematically valid? A score of 0 indicates entirely
    incorrect or absurd content; 5 indicates a fully correct solution.

    \item \textbf{Reasoning clarity} (0--5): Is the reasoning clear, logical,
    and well-structured? Are the steps easy to follow?

    \item \textbf{Pedagogical quality} (0--5): Does the answer help the
    student understand the method, not merely obtain the result? Does it
    have educational value beyond a bare computation?

    \item \textbf{Bac style adherence} (0--5): Does the answer resemble an
    official Tunisian Baccalaureate correction? Does it use the formal
    conventions (``On a\ldots'', ``Donc\ldots'', ``D'après\ldots''),
    academic French, and the expected level of detail?
\end{enumerate}

The four criteria are equally weighted. No composite score was imposed
during grading; the teacher scored each criterion independently.

\subsubsection{Blind Evaluation Protocol}

To prevent evaluator bias, we implemented a fully blind grading procedure.
For each question, the three system outputs (RAG, Prompt-Only, Hybrid) were
randomly shuffled and presented under anonymous labels (Système~X, Système~Y,
Système~Z). The shuffling was performed per question using a fixed random
seed, so the mapping between anonymous labels and real system names varied
from one question to the next. The evaluator did not know which label
corresponded to which system at any point during grading.

The anonymous answer sheets and empty grading templates were generated
programmatically from the evaluation results and provided to the evaluator
as printable documents. After grading was complete, the answer key was used
to unmask the system identities and compute per-system statistics.

\subsubsection{Evaluator}

The evaluation was conducted by a single experienced Tunisian Baccalaureate
mathematics teacher. This evaluator was selected for domain expertise: deep
familiarity with the Bac correction conventions, the official curriculum
boundaries, and the pedagogical expectations of the Section Mathématiques
program. The limitation of relying on a single evaluator is acknowledged in
Section~\ref{subsec:limitations}.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Results}
\label{subsec:results}

This section reports the outcomes of the blind evaluation. Interpretation
and explanation of these results are deferred to the Discussion
(Section~\ref{sec:discussion}).

\subsubsection{Overall System Comparison}

Table~\ref{tab:human-evaluation} presents the average scores across all 15
graded questions (categories A--D) for each criterion and system.

\begin{table}[H]
\centering
\caption{Blind human evaluation: average scores per criterion (0--5 scale),
as assessed by a Tunisian Baccalaureate mathematics teacher across 15
questions (categories A--D). Higher is better.}
\label{tab:human-evaluation}
\begin{tabular}{lrrr}
\toprule
\textbf{Criterion} & \textbf{RAG} & \textbf{Prompt-Only} & \textbf{Hybrid} \\
\midrule
Mathematical Correctness & 3.53 & 4.73 & 3.40 \\
Reasoning Clarity        & 3.47 & 4.67 & 3.53 \\
Pedagogical Quality      & 2.93 & 4.13 & 3.00 \\
Bac Style Adherence      & 2.93 & 3.73 & 3.07 \\
\midrule
\textbf{Mean (unweighted)} & \textbf{3.22} & \textbf{4.32} & \textbf{3.25} \\
\bottomrule
\end{tabular}
\end{table}

The Prompt-Only baseline received the highest scores on all four criteria.
Its overall mean (4.32/5) exceeded both the RAG system (3.22/5) and the
Hybrid engine (3.25/5) by more than one full point. The RAG and Hybrid
systems scored within 0.03 points of each other overall.

Among the four criteria, Bac Style Adherence received the lowest scores
across all three systems, with the RAG system and Hybrid scoring below 3.1.
Pedagogical Quality showed a similar pattern: only the Prompt-Only system
exceeded 4.0 on this dimension. Mathematical Correctness and Reasoning
Clarity were the two strongest criteria for all systems, with the
Prompt-Only baseline approaching near-perfect scores (4.73 and 4.67
respectively).


\subsubsection{Per-Category Breakdown}

Table~\ref{tab:per-category} shows the unweighted mean across all four
criteria, broken down by question category.

\begin{table}[H]
\centering
\caption{Per-category mean scores (unweighted average of all four criteria,
0--5 scale). Categories: A = direct Bac-style, B = novel chapter-based,
C = informal student questions, D = Derja/mixed language.}
\label{tab:per-category}
\begin{tabular}{llrrr}
\toprule
\textbf{Cat.} & \textbf{Description} & \textbf{RAG} & \textbf{Prompt-Only} & \textbf{Hybrid} \\
\midrule
A & Direct Bac-style        & 3.25 & 4.35 & 3.05 \\
B & Novel chapter-based     & 3.00 & 4.00 & 3.30 \\
C & Student informal        & 3.33 & 4.75 & 2.92 \\
D & Derja / mixed           & 3.50 & 4.38 & 4.12 \\
\bottomrule
\end{tabular}
\end{table}

Several patterns are visible. The Prompt-Only system scored highest in every
category. Its strongest performance was on Category~C (informal student
questions, 4.75/5), suggesting that the coaching mode and pedagogical prompt
components were particularly effective for conceptual, non-computational
questions.

Category~D (Derja/mixed language) produced the smallest gap between systems:
the Hybrid engine scored 4.12, within 0.26 points of the Prompt-Only
system (4.38). This was also the category where the Hybrid engine most
clearly outperformed the RAG system (4.12 vs.\ 3.50).

Category~B (novel questions) was the weakest category for the RAG system
(3.00), and the only category where the Hybrid engine (3.30) scored
noticeably above the RAG system.


\subsubsection{Curriculum Guardrail Results}

All four out-of-scope questions (Category~E) were evaluated on a binary
pass/fail basis: a system passes if it refuses to apply the forbidden method
or explicitly warns that it is outside the Bac program.

\begin{table}[H]
\centering
\caption{Curriculum guardrail results: number of out-of-scope questions
correctly refused (out of 4). A pass means the system refused the
forbidden method or explicitly flagged it as hors programme.}
\label{tab:guardrail-results}
\begin{tabular}{lr}
\toprule
\textbf{System} & \textbf{Pass Rate} \\
\midrule
RAG         & 4/4 \\
Prompt-Only & 4/4 \\
Hybrid      & 4/4 \\
\bottomrule
\end{tabular}
\end{table}

All three systems correctly refused all four out-of-scope requests. The
forbidden methods tested were L'Hôpital's rule, matrix diagonalization,
Taylor series, and Lebesgue integration. In each case, every system either
refused outright or redirected the student to the Bac-compatible alternative.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Per-Question Retrieval Breakdown}
\label{subsec:per-question-breakdown}

The aggregate comparison in Table~\ref{tab:human-evaluation} shows that the
Prompt-Only system outperformed both retrieval-based systems overall. To
investigate why, we examined the retrieval pipeline's behavior on each of the
19 evaluation queries. Table~\ref{tab:per-question-retrieval} reports the
routing case and best L2 distance for both the RAG and Hybrid engines.

\begin{table}[H]
\centering
\caption{Per-question retrieval metadata for the RAG and Hybrid systems.
All 19 queries were classified as Case~A (strong match, $d \leq 1.2$).
The RAG and Hybrid engines produced identical distances because the Hybrid
engine reuses the same retrieval pipeline.}
\label{tab:per-question-retrieval}
\begin{tabular}{llllr}
\toprule
\textbf{ID} & \textbf{Cat.} & \textbf{Chapter} & \textbf{Case} & \textbf{Best dist.} \\
\midrule
A01 & A & Nombres complexes        & A & 0.8030 \\
A02 & A & Suites réelles           & A & 0.4546 \\
A03 & A & Intégrales               & A & 0.7307 \\
A04 & A & Identité de Bézout       & A & 0.8215 \\
A05 & A & Équations différentielles & A & 0.7849 \\
\midrule
B01 & B & Fonction logarithme      & A & 0.6268 \\
B02 & B & Probabilités             & A & 0.3989 \\
B03 & B & Fonction exponentielle   & A & 0.6238 \\
B04 & B & Géométrie dans l'espace  & A & 0.5013 \\
B05 & B & Divisibilité dans $\mathbb{Z}$ & A & 0.6079 \\
\midrule
C01 & C & Continuité et limites    & A & 0.9192 \\
C02 & C & Dérivabilité             & A & 0.8064 \\
C03 & C & Primitives               & A & 0.7047 \\
\midrule
D01 & D & Suites réelles           & A & 0.8929 \\
D02 & D & Nombres complexes        & A & 0.9606 \\
\bottomrule
\end{tabular}
\end{table}

A striking feature of these results is uniformity: every single query—across
all five categories—was classified as Case~A by both engines. The best L2
distances ranged from 0.40 to 0.96, well below the strong-match threshold of
1.2. In other words, the retrieval pipeline consistently found documents that
were semantically close to every query, including the novel questions
(Category~B), the informal student queries (Category~C), and the Derja
questions (Category~D). The consequences of this uniformity are examined in
Section~\ref{subsec:retrieval-routing-discussion}.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Retrieval and Routing Analysis}
\label{subsec:retrieval-routing-analysis}

\subsubsection{Case Distribution}

Table~\ref{tab:case-distribution} summarizes the routing outcome across all
19 queries. Because every query fell into Case~A, the Hybrid engine's
distinctive behavior—Cases~B and~C, which merge retrieved context with
curriculum knowledge or fall back to pure prompt engineering—was never
activated during the evaluation.

\begin{table}[H]
\centering
\caption{Retrieval routing case distribution across all 19 evaluation queries.
Case~A: strong match ($d \leq 1.2$); Case~B: partial match ($1.2 < d \leq 1.6$);
Case~C: no useful match ($d > 1.6$).}
\label{tab:case-distribution}
\begin{tabular}{lrrl}
\toprule
\textbf{Routing Case} & \textbf{RAG} & \textbf{Hybrid} & \textbf{Description} \\
\midrule
Case~A & 19 & 19 & Strong retrieval ($d \leq 1.2$) \\
Case~B &  0 &  0 & Partial match ($1.2 < d \leq 1.6$) \\
Case~C &  0 &  0 & No useful match ($d > 1.6$) \\
\bottomrule
\end{tabular}
\end{table}

This result has an important structural implication: because the Hybrid engine
operated exclusively in Case~A throughout the evaluation, it behaved
identically to the standalone RAG system for every query. The three-case
routing mechanism—the central architectural contribution of the Hybrid
engine—was effectively inert. This explains the near-identical scores observed
in Table~\ref{tab:human-evaluation} (RAG: 3.22, Hybrid: 3.25).

\subsubsection{First-Pass vs.\ Second-Pass Retrieval}

The RAG pipeline uses a two-stage retrieval strategy: a first pass targeting
corrections (\texttt{is\_solution=true}) and a second pass targeting course
material (cours), activated only when the first pass yields a best distance
above the strong-match threshold ($d > 1.2$).

\begin{table}[H]
\centering
\caption{Two-stage retrieval behavior per graded query (categories A--D).
The first pass always retrieved 10 correction-type documents. The second
pass (course material) was never activated because all best distances fell
below 1.2. ``Selected'' is the number of documents ultimately sent to the LLM
after filtering and companion pairing.}
\label{tab:two-stage-retrieval}
\begin{tabular}{llrrrr}
\toprule
\textbf{ID} & \textbf{Case} & \textbf{Best dist.} & \textbf{1st pass} & \textbf{2nd pass} & \textbf{Selected} \\
\midrule
A01 & A & 0.80 & 10 & 0 & 7 \\
A02 & A & 0.45 & 10 & 0 & 11 \\
A03 & A & 0.73 & 10 & 0 & 11 \\
A04 & A & 0.82 & 10 & 0 & 7 \\
A05 & A & 0.78 & 10 & 0 & 12 \\
\midrule
B01 & A & 0.63 & 10 & 0 & 7 \\
B02 & A & 0.40 & 10 & 0 & 6 \\
B03 & A & 0.62 & 10 & 0 & 12 \\
B04 & A & 0.50 & 10 & 0 & 6 \\
B05 & A & 0.61 & 10 & 0 & 7 \\
\midrule
C01 & A & 0.92 & 10 & 0 & 10 \\
C02 & A & 0.81 & 10 & 0 & 13 \\
C03 & A & 0.70 & 10 & 0 & 12 \\
\midrule
D01 & A & 0.89 & 10 & 0 & 8 \\
D02 & A & 0.96 & 10 & 0 & 6 \\
\bottomrule
\end{tabular}
\end{table}

The second pass was never triggered. The first pass consistently returned 10
correction documents, and the selection stage (filtering, top-$n$ truncation,
and companion pairing) reduced these to between 6 and 13 documents per query.
The absence of course material in the retrieved context means the LLM was
conditioned exclusively on past corrections and exercise solutions, never on
theorem definitions or course notes. This is a relevant observation: for
coaching-style questions (Category~C) that ask for conceptual explanations
rather than worked solutions, correction-only context may be poorly suited.

\subsubsection{Retrieval Quality: Distance Distribution}

\begin{table}[H]
\centering
\caption{L2 distance statistics for the best-matching retrieved document
across all 19 queries. Lower distance indicates better semantic match.
All values fall well below the strong-match threshold of 1.2.}
\label{tab:distance-stats}
\begin{tabular}{lr}
\toprule
\textbf{Statistic} & \textbf{Value} \\
\midrule
Mean best distance   & 0.7361 \\
Median best distance & 0.7901 \\
Min best distance    & 0.3989 \\
Max best distance    & 0.9606 \\
Std deviation        & 0.1561 \\
\midrule
Strong-match threshold & 1.2000 \\
\bottomrule
\end{tabular}
\end{table}

The distance distribution is tightly clustered: 10 of the 19 queries had a
best distance below 0.8, and the remaining 9 fell between 0.8 and 1.0. No
query exceeded 1.0. By the L2 distance metric alone, retrieval appears
consistently strong. However, as the chapter match analysis below reveals,
low distance does not guarantee that the retrieved documents are topically
appropriate.

\subsubsection{Chapter Match Analysis}
\label{subsubsec:chapter-match}

A retrieved document is only useful if it addresses the correct mathematical
topic. We verified whether the chapter metadata of the retrieved documents
matched the expected chapter for each query.
Table~\ref{tab:chapter-match} presents the results.

\begin{table}[H]
\centering
\caption{Chapter match analysis: whether the top retrieved documents included
at least one document from the expected chapter. ``Retrieved chapters'' lists
the distinct chapters present in the selected documents (truncated to three).
Overall match rate: 8 out of 19 queries (42\%).}
\label{tab:chapter-match}
\begin{tabular}{lllll}
\toprule
\textbf{ID} & \textbf{Expected chapter} & \textbf{Retrieved chapters (sample)} & \textbf{Match} \\
\midrule
A01 & Nombres complexes     & Bac corrections, Géométrie dans l'espace & MISS \\
A02 & Suites réelles        & Intégrales, Fct.\ logarithme, Bac corrections & MISS \\
A03 & Intégrales            & Intégrales, Fct.\ exponentielle, Bac corrections & MATCH \\
A04 & Identité de Bézout    & Bac corrections, Identité de Bézout & MATCH \\
A05 & Éq.\ différentielles & Bac corrections, Primitives, Géom.\ espace & MISS \\
\midrule
B01 & Fct.\ logarithme     & Bac corrections & MISS \\
B02 & Probabilités          & Bac corrections & MISS \\
B03 & Fct.\ exponentielle  & Fct.\ exponentielle, Dérivabilité, Fct.\ logarithme & MATCH \\
B04 & Géom.\ dans l'espace & Bac corrections, Identité de Bézout & MATCH \\
B05 & Divisibilité dans $\mathbb{Z}$ & Bac corrections, Identité de Bézout & MATCH \\
\midrule
C01 & Continuité et limites & Bac corrections, Probabilités, Continuité & MATCH \\
C02 & Dérivabilité          & Fct.\ logarithme, Dérivabilité, Continuité & MATCH \\
C03 & Primitives            & Primitives & MATCH \\
\midrule
D01 & Suites réelles        & Suites réelles, Bac corrections & MATCH\tablefootnote{Category~D match rate was notably higher than Category~A, possibly because coaching-mode queries use vocabulary closer to chapter names.} \\
D02 & Nombres complexes     & Bac corrections, Bac corrections & MISS \\
\bottomrule
\end{tabular}
\end{table}

Only 8 of the 19 queries (42\%) had at least one retrieved document from
the expected chapter. The remaining 11 queries retrieved documents that were
semantically close in embedding space but addressed different mathematical
topics. A recurring pattern was the retrieval of generic ``Bac avec
corrections Images'' documents—multi-topic Baccalaureate exam compilations
that span several chapters in a single file. These documents score low L2
distances because they contain diverse mathematical content, but they may
not include the specific topic needed for a given query.

The retrieved document types confirm this pattern: of the 142 documents
selected across all 19 queries, 89 were from \textit{série} (exercise
series) collections and 53 from \textit{bac\_officiel} (official Bac exam)
compilations. No course material was retrieved, consistent with the
second-pass never being activated.

This disconnect between distance-based retrieval quality (all queries below
1.0) and topical relevance (only 42\% chapter match) is a central finding.
Its implications for the overall system comparison are examined in
Section~\ref{subsec:retrieval-routing-discussion}.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Qualitative Examples}
\label{subsec:qualitative-examples}

The aggregate scores in Table~\ref{tab:human-evaluation} show that the
Prompt-Only system outperformed the retrieval-based systems overall. To make
the differences concrete, we present excerpted outputs from three
representative queries. Each example is selected to illustrate a distinct
failure or success mode of the retrieval pipeline.

% -- Custom style for output excerpts --
\lstdefinestyle{sysoutput}{
  basicstyle=\small\ttfamily,
  breaklines=true,
  frame=single,
  numbers=none,
  xleftmargin=1em,
  xrightmargin=1em,
  aboveskip=0.5em,
  belowskip=0.5em,
}

% ·······································································
\subsubsection{Example~1: Prompt-Only Wins Despite Low Retrieval Distance
  (B02 --- Probabilités)}

\noindent\textbf{Query:} \textit{``Une urne contient 5 boules blanches
et 3 boules noires. On tire successivement 3 boules sans remise. Calculer
la probabilité d'obtenir exactement 2 boules blanches.''}

\smallskip\noindent
This query achieved the lowest retrieval distance in the entire evaluation
($d = 0.40$). Yet the chapter match was a miss: the retrieved documents were
generic ``Bac avec corrections'' compilations, not probability-specific
material.

\begin{table}[H]
\centering
\caption{Excerpted openings for query B02 (Probabilités, $d=0.40$).}
\label{tab:qual-b02}
\small
\begin{tabularx}{\textwidth}{l X}
\toprule
\textbf{System} & \textbf{Opening excerpt} \\
\midrule
RAG \newline {\scriptsize (524 chars)} &
\texttt{D'après la Source 2, on utilise $P(E) =
\frac{\text{favorable}}{\text{total}}$ \ldots}
\newline
\textit{Applies the correct formula but produces only a brief
calculation. Cites a generic retrieved source rather than
probability-specific material.} \\
\addlinespace
Prompt-Only \newline {\scriptsize (4{,}614 chars)} &
\texttt{Chapitre : Probabilités \& Dénombrement \newline
Méthode : Tirage successif sans remise $\Rightarrow$ arrangements
$A_n^p$ \ldots}
\newline
\textit{Identifies the exercise type, recalls the combinatorial formula
for draws without replacement, and works through both sub-questions
step by step with explicit intermediate calculations.} \\
\addlinespace
Hybrid \newline {\scriptsize (4{,}057 chars)} &
\texttt{D'après le contexte fourni, pour un tirage simultané de 3 boules
\ldots}
\newline
\textit{Produces a longer answer than RAG, but references the retrieved
context for ``simultaneous'' draws---a different problem type than the
sequential draws the question specifies.} \\
\bottomrule
\end{tabularx}
\end{table}

\noindent\textbf{Takeaway.}\enspace
Even with the best retrieval distance in the evaluation, the RAG system
produced a shorter and less structured answer than the Prompt-Only baseline.
The retrieved context was semantically close but addressed a variant of the
problem (simultaneous vs.\ sequential draws), and the model's instruction to
rely on provided context constrained its response. The Prompt-Only system,
unconstrained by retrieval, correctly identified the exercise type and
applied the appropriate combinatorial method.

% ·······································································
\subsubsection{Example~2: RAG Competitive --- Correct Chapter Match
  (D01 --- Suites réelles)}

\noindent\textbf{Query (Tunisian Derja):} \textit{``Kifech nethbet elli
suite monotone et bornée convergente? 3tini el méthode bil français
mathematique.''} \newline
\textit{(Translation: How do I prove that a monotone bounded sequence
converges? Give me the method in formal French mathematics.)}

\smallskip\noindent
This is one of the few queries where the chapter match succeeded
(Table~\ref{tab:chapter-match}): the retrieved documents included material
from the ``Suites réelles'' chapter.

\begin{table}[H]
\centering
\caption{Excerpted openings for query D01 (Suites réelles, $d=0.89$,
chapter match: \textsc{match}).}
\label{tab:qual-d01}
\small
\begin{tabularx}{\textwidth}{l X}
\toprule
\textbf{System} & \textbf{Opening excerpt} \\
\midrule
RAG \newline {\scriptsize (5{,}150 chars)} &
\texttt{Ahla bik! 🎓 Suites réelles --- Convergence d'une suite monotone
bornée. \newline
Théorème admis : Toute suite croissante et majorée est convergente
\ldots}
\newline
\textit{Opens with a Derja greeting, correctly identifies the chapter,
and states the monotone convergence theorem from the retrieved context.
Proceeds with a structured proof methodology.} \\
\addlinespace
Prompt-Only \newline {\scriptsize (7{,}891 chars)} &
\texttt{Chapitre : Suites Réelles (Chapitre 3) \newline
Méthode générale pour démontrer la convergence d'une suite monotone
bornée : \newline
Étape 1 : Montrer que $(u_n)$ est monotone \ldots}
\newline
\textit{Produces the longest answer: comprehensive treatment with explicit
step-by-step methodology and worked example. No Derja greeting.} \\
\addlinespace
Hybrid \newline {\scriptsize (4{,}661 chars)} &
\texttt{Ahla! Suites réelles --- Théorème de convergence monotone. \newline
D'après le contexte, le théorème fondamental est \ldots}
\newline
\textit{Also identifies the chapter correctly and cites the theorem.
More concise than Prompt-Only but focused and accurate.} \\
\bottomrule
\end{tabularx}
\end{table}

\noindent\textbf{Takeaway.}\enspace
When the retrieval pipeline returns topically relevant content---as it did
here with a correct chapter match---all three systems produce substantive
answers. The Prompt-Only system still generated a longer and more
comprehensive response, but the quality gap was visibly narrower than in
Example~1. This pattern is consistent with the per-category results in
Table~\ref{tab:per-category}, where Category~D showed the smallest
inter-system gap (RAG: 3.50, Hybrid: 4.12, Prompt-Only: 4.38). It suggests
that corpus coverage, not embedding quality, is the binding constraint:
when retrieval finds the right chapter, the retrieval-based systems approach
the performance of the unconstrained baseline.

% ·······································································
\subsubsection{Example~3: Context Mismatch on a Coaching Question
  (C01 --- Continuité)}

\noindent\textbf{Query:} \textit{``Je comprends pas comment on étudie la
continuité d'une fonction en un point. C'est quoi la méthode? Donne moi un
exemple simple svp.''}

\smallskip\noindent
The student asks for a definition and a simple example of
\textit{pointwise} continuity. The retrieval pipeline found documents about
continuity \textit{on intervals}---a related but distinct topic ($d=0.92$,
chapter match: \textsc{match}, but wrong subtopic).

\begin{table}[H]
\centering
\caption{Excerpted openings for query C01 (Continuité, $d=0.92$).}
\label{tab:qual-c01}
\small
\begin{tabularx}{\textwidth}{l X}
\toprule
\textbf{System} & \textbf{Opening excerpt} \\
\midrule
RAG \newline {\scriptsize (3{,}014 chars)} &
\texttt{Le CONTEXTE fourni contient des exemples de démonstration de la
continuité de fonctions sur des intervalles \ldots}
\newline
\textit{Explicitly acknowledges that the retrieved context addresses
interval continuity. Adapts this context but never provides the three
conditions for pointwise continuity that the student asked for.} \\
\addlinespace
Prompt-Only \newline {\scriptsize (5{,}019 chars)} &
\texttt{Chapitre : Fonctions Numériques (Chapitre 2) \newline
La continuité en un point $a$ : $f$ est continue en $a$ si et seulement si
$\lim_{x \to a} f(x) = f(a)$ \ldots}
\newline
\textit{States the formal definition directly, then works through a
concrete example ($f(x) = x^2$ at $x = 1$) as the student requested.} \\
\addlinespace
Hybrid \newline {\scriptsize (3{,}865 chars)} &
\texttt{D'après le contexte, nous avons des exemples de continuité de
fonctions sur un intervalle. Il ne contient pas la définition formelle ni
la méthode pour un point \ldots}
\newline
\textit{Recognizes the mismatch between the retrieved context and the
question, yet remains constrained by its Case~A instructions to treat the
context as primary. Provides a partial answer.} \\
\bottomrule
\end{tabularx}
\end{table}

\noindent\textbf{Takeaway.}\enspace
This example illustrates the subtopic mismatch problem: the retrieval
distance was reasonable, the chapter was technically correct (both topics
fall under ``Continuité et limites''), but the retrieved content addressed
the wrong aspect of continuity. The Prompt-Only system, free to draw on its
parametric knowledge, directly provided the definition and example the
student needed. The Hybrid engine's acknowledgment that the context ``ne
contient pas la définition formelle'' is particularly revealing: the system
detected the mismatch but could not override its retrieval-first
instructions in Case~A. This observation motivates the threshold
recalibration discussed in Section~\ref{sec:future-work}.

\paragraph{Summary of qualitative patterns.}
Across the three examples, a consistent pattern emerges. The Prompt-Only
system produces longer, more structured, and more pedagogically complete
answers by drawing freely on its parametric knowledge. The retrieval-based
systems (RAG and Hybrid) are constrained by the quality and topical
relevance of the retrieved context: when retrieval finds the correct chapter
(Example~2), the gap narrows; when retrieval returns semantically close but
topically mismatched content (Examples~1 and~3), the constraint actively
harms output quality. These qualitative observations are consistent with the
quantitative finding that corpus coverage---not embedding distance---is the
primary bottleneck.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Embedding Model Comparison: BGE-M3 vs.\ Google text-embedding}
\label{subsec:embedding-comparison}

The choice of embedding model directly affects retrieval quality. The
production system uses BGE-M3~\cite{chen2024bge_m3}, a multilingual model
with 1024-dimensional dense vectors. To evaluate whether an alternative
embedding model would yield better retrieval, we compared BGE-M3 against
Google's \texttt{text-embedding-005} model (768-dimensional vectors) available
through the Vertex AI API.

\subsubsection{Experimental Setup}

Both models were used to embed the same 15 graded evaluation queries and the
full corpus of chunks. For each query, we retrieved the top-10 nearest
documents using L2 distance and compared the results along three dimensions:
\begin{enumerate}
    \item \textbf{Best L2 distance}: How close is the nearest document to the
    query? Lower distances indicate stronger semantic matches.
    \item \textbf{Routing case distribution}: How do the distances map to the
    three routing cases (A/B/C) used by the Hybrid engine?
    \item \textbf{Top-5 document overlap}: Do both models retrieve the same
    documents, or do they find fundamentally different content?
\end{enumerate}

\subsubsection{Results}

\begin{table}[H]
\centering
\caption{Aggregate comparison of BGE-M3 vs.\ Google text-embedding-005 for
retrieval on the Tunisian Bac Math corpus (15 queries, 903 corpus chunks).
L2 distance metrics are computed from the best-matching document per query.}
\label{tab:embedding-aggregate}
\begin{tabular}{lrr}
\toprule
\textbf{Metric} & \textbf{BGE-M3} & \textbf{Google} \\
\midrule
Embedding dimension      & 1024 & 768 \\
Mean best L2 distance    & 0.6879 & 0.5116 \\
Median best L2 distance  & 0.7307 & 0.5226 \\
Queries in Case~A        & 15 & 15 \\
Queries in Case~B        & 0 & 0 \\
Queries in Case~C        & 0 & 0 \\
Mean top-5 overlap ratio & \multicolumn{2}{c}{0.23} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[H]
\centering
\caption{Per-query best L2 distance and routing case for BGE-M3 vs.\ Google
text-embedding-005. Lower distance = better retrieval match.
Cases: A = strong ($\leq 1.2$), B = weak ($\leq 1.6$), C = no match ($> 1.6$).}
\label{tab:embedding-per-query}
\begin{tabular}{llrrlrl}
\toprule
\textbf{ID} & \textbf{Cat.} & \multicolumn{2}{c}{\textbf{BGE-M3}} & &
\multicolumn{2}{c}{\textbf{Google}} \\
\cmidrule(lr){3-4} \cmidrule(lr){6-7}
& & \textbf{Dist.} & \textbf{Case} & & \textbf{Dist.} & \textbf{Case} \\
\midrule
A01 & A & 0.7623 & A & & 0.4783 & A \\
A02 & A & 0.3332 & A & & 0.4357 & A \\
A03 & A & 0.7307 & A & & 0.4763 & A \\
A04 & A & 0.8215 & A & & 0.6138 & A \\
A05 & A & 0.7849 & A & & 0.6494 & A \\
B01 & B & 0.6268 & A & & 0.4537 & A \\
B02 & B & 0.3989 & A & & 0.3244 & A \\
B03 & B & 0.6238 & A & & 0.3911 & A \\
B04 & B & 0.5013 & A & & 0.4467 & A \\
B05 & B & 0.6079 & A & & 0.5226 & A \\
C01 & C & 0.8744 & A & & 0.5583 & A \\
C02 & C & 0.8064 & A & & 0.5708 & A \\
C03 & C & 0.7047 & A & & 0.6034 & A \\
D01 & D & 0.8113 & A & & 0.5437 & A \\
D02 & D & 0.9297 & A & & 0.6065 & A \\
\midrule
\textbf{Mean} & & \textbf{0.6879} & & & \textbf{0.5116} & \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Observations}

Five findings emerge from the embedding comparison experiment.

\paragraph{Google produces uniformly lower distances.}
The Google model achieves a mean best L2 distance of 0.5116, compared to
0.6879 for BGE-M3—a 25.6\% reduction (Table~\ref{tab:embedding-aggregate}).
This advantage is consistent: Google yields a lower best distance on 14 of
15 queries (Table~\ref{tab:embedding-per-query}), with query A02 being the
single exception (BGE-M3: 0.3332, Google: 0.4357). The lower standard
deviation for Google (0.0890 vs.\ 0.1665) indicates more uniform retrieval
quality across queries.

\paragraph{Both models route all queries to Case~A.}
Despite the distance difference, both models place all 15 queries in Case~A
(best distance $\leq 1.2$), meaning the retrieval routing thresholds used by
the Hybrid engine would produce identical routing decisions regardless of
which embedding model is deployed. No query falls into Case~B or Case~C for
either model.

\paragraph{Retrieved documents differ substantially.}
The mean top-5 overlap ratio of 0.23 indicates that only approximately one
in five documents appears in the top-5 results of both models. This low
overlap demonstrates that the two models construct meaningfully different
semantic representations of mathematical content and retrieve different
corpus chunks for the same query.

\paragraph{Category-level patterns.}
The per-category means reveal that Google's advantage is largest for
Category~B questions (BGE-M3: 0.5517, Google: 0.4277—a 22.5\% reduction)
and Category~C questions (BGE-M3: 0.7952, Google: 0.5775—a 27.4\%
reduction). Categories~C and~D, which contain more complex multi-part
questions, show the highest absolute distances for both models, suggesting
that these queries are inherently harder to match against the corpus.

\paragraph{Implications for the production system.}
The results suggest that Google's \texttt{text-embedding-005} produces
tighter semantic matches on this corpus. However, three factors justify
retaining BGE-M3 in the production pipeline. First, the routing outcomes are
identical—both models assign Case~A to every query—so the Hybrid engine
would make the same retrieval-vs-generation decision with either model.
Second, BGE-M3 is a local model that requires no API calls, eliminating
latency, cost, and rate-limit concerns at inference time. Third, given that
the blind evaluation (Section~\ref{subsec:blind-evaluation}) showed that the
Prompt-Only baseline outperformed the retrieval-based systems overall, the
bottleneck in system performance lies in corpus coverage and retrieval
integration rather than in embedding quality. Switching to Google embeddings
would produce lower distances but would not address the underlying corpus
completeness issues identified in Section~\ref{subsec:interpretation}.


% ─────────────────────────────────────────────────────────────────────────────
\subsection{Digitization Quality Evaluation}
\label{subsec:ocr-evaluation}

The quality of the RAG system's knowledge base depends directly on the
accuracy of the digitization pipeline described in
Section~\ref{subsec:digitization}. Errors introduced during OCR—a misread
exponent, a dropped minus sign, a garbled fraction—propagate silently into
the vector store and can corrupt the retrieved context presented to the
generative model. To quantify this risk, we conducted a ground truth
evaluation of the Gemini multimodal OCR output against manually corrected
reference texts.

\subsubsection{Metric Selection}

We report three metrics, each capturing a different dimension of
digitization fidelity.

\paragraph{Character Error Rate (CER).}
CER is the standard metric for OCR evaluation~\cite{wang2023ocrbench}. It
is defined as the character-level Levenshtein edit distance between the OCR
output and the reference text, normalized by the length of the reference:
\[
\text{CER} = \frac{\text{editDistance}(\text{OCR}, \text{ref})}{|\text{ref}|}
\]
A CER of 0\% indicates a perfect transcription; a CER of 5\% means that
roughly one in twenty characters requires correction. CER captures all
errors equally—whitespace, punctuation, letters, and mathematical
symbols—without distinguishing their semantic impact.

We chose CER because it is the most widely used metric in the OCR
literature and allows comparison with published
benchmarks~\cite{wang2023ocrbench, mahdavi2019icdar_math}. Its main
limitation is that it treats all character errors equally: a misplaced
space and a flipped sign in an exponent count the same, even though the
latter is mathematically catastrophic.

\paragraph{Word Error Rate (WER).}
WER operates at the word level, computing the Levenshtein distance between
the tokenized OCR output and the tokenized reference, normalized by the
reference word count. WER penalizes word-level substitutions, insertions,
and deletions, and is less sensitive to minor formatting differences
(extra spaces, line break variations) than CER. It is commonly used in
speech recognition evaluation and has been adopted for OCR evaluation of
structured documents.

\paragraph{\LaTeX{} Math Expression Accuracy.}
Neither CER nor WER distinguishes mathematical content from natural
language text. A document could have perfect CER on the French prose but
corrupt every equation, or vice versa. To address this, we compute a
\LaTeX{}-specific metric: the fraction of math expressions in the
reference text (delimited by \texttt{\$...\$}, \texttt{\$\$...\$\$},
\texttt{\textbackslash(...\textbackslash)}, or
\texttt{\textbackslash[...\textbackslash]}) that appear exactly in the OCR
output. This metric directly measures what matters most for a mathematical
RAG system: whether the equations that will be embedded and retrieved are
faithful to the source material.

This metric is an implementation-specific design. We are not aware of a
published OCR metric that specifically targets \LaTeX{} math expression
fidelity, though related work on mathematical expression
recognition~\cite{mahdavi2019icdar_math} evaluates similar properties at
the symbol and structure level.


\subsubsection{Ground Truth Methodology}

We selected a stratified sample of digitized documents from the corpus,
aiming to cover the range of document types and difficulty levels present
in the knowledge base:

\begin{itemize}
    \item Corrections with dense mathematical notation (complex numbers,
    integrals)
    \item Corrections with text-heavy reasoning (proofs, sequence
    convergence)
    \item Course material with theorem statements and definitions
    \item Documents with fractions, summation notation, and multi-line
    equations (high OCR difficulty)
    \item Cleanly printed documents (expected low error rate, serving as
    a baseline)
\end{itemize}

For each selected document, the Gemini OCR output (\texttt{.tex} file
from GCS) was paired with a manually corrected reference version. The
manual correction was performed by the author, consulting the original
scanned image to resolve any ambiguities. The corrected reference preserves
the same \LaTeX{} formatting conventions as the OCR output so that the
comparison measures transcription accuracy rather than stylistic
preferences.

% ── RESULTS TABLE ──────────────────────────────────────────────────────────

\subsubsection{Results}

Table~\ref{tab:ocr-quality} presents the per-sample and aggregate results.

\begin{table}[H]
\centering
\caption{OCR digitization quality: Character Error Rate (CER), Word Error
Rate (WER), and \LaTeX{} math expression accuracy, measured on a stratified
sample of five ground truth pairs. Lower is better for CER/WER; higher is
better for \LaTeX{} accuracy. Sample~3 contained no inline math delimiters
in the OCR output, so \LaTeX{} accuracy is not applicable.}
\label{tab:ocr-quality}
\begin{tabular}{llrrr}
\toprule
\textbf{Sample} & \textbf{Doc.\ type} & \textbf{CER (\%)} & \textbf{WER (\%)} & \textbf{\LaTeX{} Acc.\ (\%)} \\
\midrule
Sample~1 & Correction (complexes)     & 5.4  & 10.0 & 66.7 \\
Sample~2 & Cours (suites)             & 4.2  & 5.5  & 99.0 \\
Sample~3 & Correction (intégrales)    & 0.0  & 0.0  & ---  \\
Sample~4 & Correction (probabilités)  & 4.1  & 5.8  & 100.0 \\
Sample~5 & Cours (théorèmes)          & 13.4 & 8.8  & 95.5 \\
\midrule
\textbf{Average} & --- & \textbf{5.4} & \textbf{6.0} & \textbf{97.1} \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Observations}

The average CER of 5.4\% and WER of 6.0\% indicate that the Gemini
multimodal OCR pipeline produces largely faithful transcriptions of the
source material. Four of the five samples fall below 6\% on both metrics,
which is consistent with the performance ranges reported for multimodal LLM
OCR on structured documents~\cite{wang2023ocrbench}. The near-perfect
scores on Sample~3 (CER and WER both 0.0\%) suggest that cleanly printed
correction pages with standard formatting pose little difficulty for the
model.

Sample~5 (cours, théorèmes) was the weakest performer at 13.4\% CER, which
is notably higher than the other samples. This document contained dense
theorem statements with nested mathematical environments and multi-line
display equations. The higher error rate likely reflects the difficulty of
parsing complex \LaTeX{} structures from scanned images, particularly when
line breaks, alignment environments, and nested delimiters are involved.

The \LaTeX{} math expression accuracy averaged 97.1\% across the four
samples where it could be measured, meaning that the vast majority of
mathematical expressions were transcribed without alteration. Sample~1
(complexes correction) was the exception, with only 2 of 3 math expressions
preserved exactly (66.7\%). While this sample size is too small to draw
strong conclusions from, it suggests that isolated errors in mathematical
notation do occur and can affect the correctness of retrieved context in the
RAG pipeline.

These results have two implications for the system evaluation. First, the
overall digitization quality is high enough that the knowledge base can be
considered a reasonably faithful representation of the source material—the
corpus is not dominated by OCR noise. Second, the occasional errors that do
occur (roughly 1 in 20 characters, and 1 in 30 math expressions) could
contribute to the lower performance of the RAG and Hybrid systems observed
in the blind evaluation (Section~\ref{subsec:results}): when the retrieved
context contains a corrupted equation or a garbled reasoning step, the
generative model must work around it, which may degrade answer quality. The
Prompt-Only system, which never accesses the corpus, is entirely immune to
this source of error.
