\chapter{Methods}
This section describes the design and implementation of the three system
variants developed in this thesis: the RAG system
(Section~\ref{sec:rag-system}), the prompt-only baseline
(Section~\ref{sec:prompt-only}), and the hybrid engine
(Section~\ref{sec:hybrid}). It begins with a short overview of the overall
pipeline and then presents the main components in more detail in the
following subsections.

\subsection{System Overview}
\label{subsec:system-overview}

The core system developed in this thesis follows a retrieval-augmented
generation setup in which a language model is combined with a curated external
corpus. Instead of relying only on the model's parametric knowledge, the
system retrieves relevant mathematical material at inference time and provides
it as context for answer generation. This is particularly useful in the
present setting, where the goal is not only to produce correct answers, but to
generate solutions that resemble Tunisian Baccalaureate correction style.

Figure~\ref{fig:system-architecture} shows the overall pipeline. The process
starts with scanned educational material collected from Tunisian Bac
mathematics sources. These documents are digitized into \LaTeX, normalized,
chunked, embedded, and stored in a local vector database. At query time, a
student question is used to retrieve relevant material from this indexed
corpus, and the retrieved context is then passed to the generation model.

The implementation can be described in five stages: data acquisition,
digitization, indexing, retrieval, and generation. 

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    node distance=1.2cm and 1.8cm,
    stage/.style={rectangle, draw=black!70, fill=blue!8, rounded corners=4pt,
                  minimum width=3.2cm, minimum height=1.1cm, align=center,
                  font=\small\bfseries},
    comp/.style={rectangle, draw=black!50, fill=white, rounded corners=2pt,
                 minimum width=2.6cm, minimum height=0.8cm, align=center,
                 font=\small},
    arrow/.style={->, >=stealth, thick, draw=black!60},
    label/.style={font=\footnotesize\itshape, text=black!60},
]

\node[stage] (acq) {Stage 1\\Data Acquisition};
\node[comp, below=0.5cm of acq] (scan) {Scanned exams,\\series, cours};

\node[stage, right=2cm of acq] (dig) {Stage 2\\Digitization};
\node[comp, below=0.5cm of dig] (gemini_ocr) {Gemini multimodal\\OCR $\to$ \LaTeX{}};

\node[stage, right=2cm of dig] (idx) {Stage 3\\Indexing};
\node[comp, below=0.5cm of idx] (chunk) {Chunking + metadata\\+ embeddings};
\node[comp, below=0.4cm of chunk] (chroma) {ChromaDB\\vector store};

\node[stage, right=2cm of idx] (ret) {Stage 4\\Retrieval};
\node[comp, below=0.5cm of ret] (twostage) {Two-stage retrieval\\+ companion fetch};

\node[stage, right=2cm of ret] (gen) {Stage 5\\Generation};
\node[comp, below=0.5cm of gen] (llm) {Gemini LLM};

\draw[arrow] (acq) -- node[above, label] {images / PDF} (dig);
\draw[arrow] (dig) -- node[above, label] {\texttt{.tex} files} (idx);
\draw[arrow] (idx) -- node[above, label] {vectors} (ret);
\draw[arrow] (ret) -- node[above, label] {context} (gen);

\draw[arrow] (scan) -- (acq);
\draw[arrow] (gemini_ocr) -- (dig);
\draw[arrow] (chunk) -- (idx);
\draw[arrow] (chroma) -- (chunk);
\draw[arrow] (twostage) -- (ret);
\draw[arrow] (llm) -- (gen);

\node[comp, above=0.8cm of ret, fill=green!10, draw=green!50!black]
    (query) {Student query};
\draw[arrow, draw=green!50!black] (query) -- (ret);

\node[comp, above=0.8cm of gen, fill=red!8, draw=red!50!black]
    (answer) {Generated answer};
\draw[arrow, draw=red!50!black] (gen) -- (answer);

\draw[dashed, gray] ([xshift=0.9cm]idx.east |- acq.north) ++(0,1.2)
    -- ++(0,-5.5) node[below, font=\footnotesize\bfseries, text=gray]
    {offline $\mid$ online};

\end{tikzpicture}%
}
\caption{Overview of the end-to-end pipeline used in the RAG system. }
\label{fig:system-architecture}
\end{figure}

\subsubsection{Pipeline Stages}

The main stages of the pipeline are briefly summarized here before being
described in detail in the following subsections.

\paragraph{Stage 1: Data Acquisition.}
The corpus was assembled from three main categories of Tunisian Baccalaureate
mathematics material: official past examinations with corrections, chapterwise
exercise series with worked solutions, and course material. A central design
choice was to preserve the link between an exercise and its correction
whenever possible, since this pairing later proved important for retrieval.

\paragraph{Stage 2: Digitization.}
Because the source documents were available mainly as scanned pages or images, they first had to be converted into machine-readable text. For this
purpose, Gemini was used as a multimodal digitization model to transcribe both
mathematical notation and surrounding natural-language text into \LaTeX.

\paragraph{Stage 3: Indexing.}
The digitized files were normalized, split into chunks, enriched with
metadata, embedded using BGE-M3, and stored in ChromaDB. This stage created
the searchable vector store used later during inference.

\paragraph{Stage 4: Retrieval.}
When a student submits a question, the system retrieves relevant material from
the vector database. The retrieval process follows a two-stage strategy: it
first searches for worked corrections and then, when needed, supplements these
with course material. If a correction is selected, the corresponding exercise
statement is also fetched whenever possible.

\paragraph{Stage 5: Generation.}
The retrieved material is inserted into a structured prompt together with the
student question and then passed to the Gemini generation model. The aim is to
produce an answer that is not only mathematically relevant, but also aligned
with the style and reasoning conventions of Tunisian Bac corrections.

\subsubsection{Shared Infrastructure Across System Variants}

All three system variants evaluated in this thesis share the same basic
infrastructure. They use the same generation model family, the same overall
user interface, and the same general stylistic constraints for answer
generation. The main difference lies in how each system constructs the context
that is given to the model.

In the RAG system, the prompt includes retrieved documents from the indexed
corpus. In the prompt-only baseline, no retrieval is performed and the prompt
itself must carry the full task guidance. In the hybrid system, the choice
between retrieval-based and prompt-only behavior depends on the quality of the
retrieved material.

This shared setup was intentional. It makes the comparison between the three
variants easier to interpret, since the main changing factor is the use of
retrieval rather than unrelated differences in interface or generation model
configuration.

\section{Retrieval-Augmented Generation System}
\label{sec:rag-system}

This section describes the design and implementation of the
retrieval-augmented generation (RAG) system developed in this thesis. In
contrast to the prompt-only baseline presented in
Section~\ref{sec:prompt-only}, this system uses retrieved documents from a
curated corpus of Tunisian Baccalaureate mathematics material as additional
context for answer generation. The implementation can be divided into six main
parts: data acquisition, mathematical document digitization, database
construction and indexing, embedding model selection, retrieval, and answer
generation. Each of these components is presented in the following
subsections, together with the main design choices that shaped the final
system.

\subsection{Data Acquisition and Paired Exercise--Correction Structure}
\label{subsec:data-acquisition}

The effectiveness of a retrieval-augmented system depends not only on the
retrieval model, but also on the quality and internal organization of the
corpus from which documents are retrieved~\cite{lewis2020rag,gao2024rag_survey}.
For this reason, corpus construction was treated as a central part of the
system design rather than as a simple preprocessing step. In building the
dataset, priority was given to source quality, structural consistency, and the
preservation of pedagogically meaningful document units.

The target domain of the present work is Tunisian Baccalaureate mathematics
(Section Mathématiques). This examination context is highly standardized, not
only in terms of mathematical content but also in terms of redaction style.
Successful answers are expected to follow recognizable conventions of
presentation, justification, and logical sequencing. However, no publicly
available machine-readable corpus of such material was available for direct
use in this project. The corpus therefore had to be assembled manually from
physical and scanned educational sources collected in Tunisia.

\subsubsection{Physical Data Collection}


The raw material was collected during a dedicated trip to Tunisia. The main
sources were books and course materials prepared by experienced teachers who
specialize in the Tunisian Baccalaureate mathematics program (Section
Mathématiques). These materials are widely used by students preparing for the
exam and typically contain both exercises and fully worked corrections written
in the formal style expected in Bac examinations.

Preparing this material for digitization raised two practical difficulties.
Printed pages, such as typed exercise sheets or course notes, could usually be
scanned without major problems using either a flatbed scanner or a
high-resolution phone camera. Handwritten corrections were more difficult.
Some were still legible enough to be scanned directly, but others contained
dense notation, small handwriting, or faded ink and could not be converted
reliably by OCR tools, including multimodal systems designed for visually
complex documents~\cite{wang2023ocrbench, mahdavi2019icdar_math}. In those
cases, the handwritten page was rewritten manually on a tablet before being
scanned again. Although this added a manual step to the workflow, it helped
avoid introducing low-quality source material into the retrieval corpus and
was therefore consistent with the emphasis on upstream data quality discussed
above~\cite{sambasivan2021everyone}.

A further decision concerned the unit of scanning. Rather than digitizing
entire exercise sheets and their corrections as large undivided documents, the
material was prepared at the level of individual exercises whenever possible.
In practice, this meant preserving the link between a given exercise and its
corresponding correction as closely as the source material allowed. For course
documents, by contrast, chapter material was typically scanned as a continuous
document. This distinction was intentional and later became important for the
retrieval stage.

\subsubsection{Corpus Structure and File Organization}

After scanning, the material was uploaded to a Google Cloud Storage (GCS)
bucket~\cite{gcloud_storage2024}. The corpus was organized in a hierarchical
folder structure designed to reflect both the mathematical chapter and the
document type:

\begin{verbatim}
BacMath_Raw_Data/
  01_Continuite_et_limites/
    Cours/                          → course material
    series_et_corrections/          → exercises + corrections
  02_Derivabilite/
    Cours/
    series_et_corrections/
  ...
  20_Coniques/
    Cours/
    series_et_corrections/
  Bac_avec_corrections_Images/      → past exam corrections
\end{verbatim}

The resulting corpus consists of three main categories of material. The first
is chapter-based course material (\textit{Cours}), containing definitions,
theorems, and other chapter-specific content. The second is chapter-based
exercise series with worked corrections (\textit{series et corrections}).
The third is a separate collection of older official Baccalaureate
examinations together with their corrections.

This folder structure was not chosen only for convenience. It also served as a
machine-readable source of metadata. During indexing
(Section~\ref{subsec:database-construction}), the system extracts information
such as document type, chapter, year, session, and exercise identifier
directly from the folder path and file name. For example, a path such as
\texttt{BacMath\_Raw\_Data/10\_Nombres\_complexes/series\_et\_corrections/S1\_EX1\_sol\_1.jpg.tex}
can be parsed automatically to recover the chapter (\textit{Nombres
complexes}), the document type (\textit{serie}), the exercise key
(\texttt{S1\_EX1}), and the fact that the file corresponds to a solution.
This made it possible to attach structured metadata to indexed chunks without
manual annotation.

The corpus structure was also designed to remain easy to extend. Adding new
material does not require redesigning the pipeline; it only requires placing
the files in the appropriate chapter folder and subfolder. In its current
state, the corpus spans 20 chapters and contains 624 indexed source files,
which are represented as 1{,}422 retrieval chunks after indexing. A summary of
the indexed corpus is given in Table~\ref{tab:corpus-stats}.

\begin{table}[H]
\centering
\caption{Summary of the indexed corpus.}
\label{tab:corpus-stats}
\begin{tabular}{|l|r|}
\hline
\textbf{Item} & \textbf{Count} \\
\hline
Chapters covered & 20 \\
Total indexed files & 624 \\
Total chunks & 1{,}422 \\
\hline
\textit{bac\_officiel} files & 312 \\
\textit{serie} files & 250 \\
\textit{cours} files & 62 \\
\hline
\end{tabular}
\end{table}

\subsubsection{Paired Exercise--Correction Acquisition}

A deliberate design decision was made during corpus preparation: whenever
possible, each exercise was acquired together with its corresponding
correction, rather than treating exercise statements and solutions as fully
independent documents. In this thesis, this is referred to as
\textit{paired exercise--correction acquisition}.

This choice was motivated by the nature of mathematical solutions
themselves. A worked correction is not simply a final answer; it is a chain of
reasoning whose meaning depends on the precise wording, assumptions, and
constraints of the exercise. If the statement and the correction are detached
from one another too early in the pipeline, part of that logical connection is
lost. From a retrieval perspective, this is also undesirable: when a student
asks a question, the most useful reference is usually a complete worked
example, not a solution presented without its original problem statement.

The decision was also consistent with broader findings on chunking and
document structure in retrieval systems. Prior work suggests that retrieval
often improves when document units follow meaningful internal boundaries rather
than arbitrary character limits. Zhao~et~al.~\cite{zhao2024financial_chunking}
show that structure-aware chunking can improve retrieval quality, and Chen
et~al.~\cite{chen2024mog} likewise argue that different granularities may be
appropriate for different retrieval needs. Although these studies do not focus
specifically on mathematical exercises in the Tunisian Baccalaureate context,
the same general principle applies here: in the present corpus, the exercise
together with its correction forms a natural pedagogical unit, and preserving
that unit during acquisition made the later retrieval stage more coherent.

No prior work has studied paired exercise--correction acquisition for
mathematical RAG systems in the specific context of the Tunisian
Baccalaureate. This design decision is therefore presented as a
domain-specific implementation choice of this thesis, grounded in broader
principles of structure-aware document processing for retrieval
systems~\cite{zhao2024financial_chunking, chen2024mog}.

\subsection{Mathematical Document Digitization}
\label{subsec:digitization}

Once the raw material had been scanned, it still had to be converted into a
machine-readable form suitable for indexing and retrieval. This was not a
trivial step. Mathematical documents contain expressions whose meaning depends
on spatial structure, such as fractions, superscripts, subscripts, integrals,
or matrices, and these are not handled well by OCR systems designed primarily
for linear text~\cite{wang2023ocrbench, mahdavi2019icdar_math}. For the
present project, the digitization stage therefore had to preserve both the
surrounding French text and the mathematical notation in a format that could
later be embedded and retrieved. \LaTeX{} was chosen for this purpose because
it provides a structured textual representation of mathematical expressions.

\subsubsection{Tool Selection: Mathpix versus Gemini}

At an early stage, we considered using \textbf{Mathpix}, a commercial tool
specialized in mathematical OCR and equation-to-\LaTeX{} conversion~\cite{mathpix2024}.
From a purely technical perspective, this would have been a reasonable choice.
However, Mathpix is billed on a per-page basis, and the amount of scanned
material collected for this thesis made that option difficult to justify
within the available resources.

For that reason, the digitization pipeline was built around \textbf{Google
Gemini}, accessed through Vertex AI. Gemini is not a dedicated math-OCR tool,
but it is a multimodal model that can process scanned images and produce
structured text output. Recent evaluations suggest that large multimodal
models can perform competitively on OCR-like tasks involving visually complex
documents, including mathematical expressions, while still remaining sensitive
to handwriting quality and scan degradation~\cite{wang2023ocrbench}. In our
case, the corpus consisted mainly of printed material and carefully prepared
handwritten pages, so Gemini proved sufficiently usable for the task.

This choice should not be interpreted as a claim that Gemini is generally
superior to specialized mathematical OCR tools. Rather, it was a practical
decision shaped by cost and infrastructure. At the time of implementation, no
direct peer-reviewed comparison between Gemini and Mathpix for this exact use
case was available. The decision to use Gemini should therefore be understood
as an implementation choice under resource constraints, not as a general
benchmark conclusion.

\subsubsection{Digitization Pipeline (\texttt{digitize.py})}

The digitization workflow was implemented in the \texttt{digitize.py} module.
Its role is to process scanned files stored in Google Cloud Storage and to
generate companion \LaTeX{} files that can later be indexed by the retrieval
pipeline.

The process begins by scanning the bucket under the project prefix
\texttt{BacMath\_Raw\_Data/} and identifying supported source files, including
PDF, PNG, JPG, JPEG, and WEBP documents. Files that correspond to virtual
environment folders, cache folders, or other irrelevant paths are ignored.

For each raw file, the script then checks whether a corresponding
\texttt{.tex} file already exists. This was important because digitization was
performed incrementally over time rather than in a single batch. If a source
file had already been processed, it was skipped. In practice, this made the
pipeline resume-friendly: interrupted runs could be restarted without
reprocessing the entire corpus.

Files that still required digitization were sent to Gemini together with a
structured prompt asking for a faithful transcription into \LaTeX{}. The prompt
instructed the model to preserve both the mathematical content and the style
of the original document as closely as possible, to keep the order of the
document intact, and to mark unreadable passages explicitly with
\texttt{[ILLISIBLE]}. It also requested a consistent document structure,
including separate headings for exercise statements and corrections when both
were present.

Once generated, the resulting \LaTeX{} text was uploaded back to the same
storage location as a companion \texttt{.tex} file. Errors were written to a
JSONL report so that failed pages could later be inspected and retried
without repeating successful transcriptions. The script also included retry
logic with exponential backoff in order to handle temporary API failures more
robustly during longer runs.

\subsubsection{Bug Discovery and Fix}

During development, a problem was discovered in an earlier version of the
deduplication logic. The script checked only for companion files of the form
\texttt{image.png.tex}, but some older transcriptions in the storage bucket
had been saved under the shorter form \texttt{image.tex}. As a result, already
digitized files could be treated as if they were still pending, which led to
unnecessary reprocessing.

This issue was corrected by making the deduplication step recognize both
naming conventions when checking whether a raw file had already been
transcribed. New uploads were then standardized under a consistent naming
scheme. Although this was a relatively small implementation bug, it had
practical consequences for cost, runtime, and reproducibility, and it
therefore mattered for the stability of the overall pipeline.

\subsubsection{Digitization Quality and OCR Groundtruth}

Digitization quality was particularly important in this project because all
later stages operate on the transcribed \LaTeX{} rather than on the original
images. Any error introduced during OCR, such as an incorrect formula, a
missing symbol, or a distorted theorem statement, becomes part of the indexed
corpus and may later be retrieved as context for answer generation. In that
sense, digitization errors do not remain local; they propagate through the
rest of the system~\cite{sambasivan2021everyone}.

For that reason, digitization quality was later examined through a small
groundtruth evaluation in which manually checked pages were compared with the
Gemini transcriptions. Since this constitutes an empirical result rather than
a design choice, the methodology and findings are reported in the Results
section. What is relevant here is that the need for such an evaluation was
already clear at the design stage, because the digitization pipeline forms the
basis of everything that follows in the RAG system.

In practical terms, the experience gained during corpus preparation confirmed
the same point. Printed pages and clean rewritten material were generally
usable after transcription, whereas degraded handwritten pages were much more
error-prone. This was one of the main reasons why the earlier manual rewriting
step was not treated as optional convenience work, but as part of corpus
quality control.

\subsection{Database Construction and Indexing}
\label{subsec:database-construction}

With the corpus digitized into \LaTeX{} files stored in GCS, the next step is
to transform these files into an indexed, searchable vector database. This is
handled by the \texttt{build\_db.py} module, which performs normalization,
chunking, metadata extraction, embedding, and storage.

\subsubsection{From JSON Vector Store to ChromaDB}

In an early prototype of the system, document embeddings were stored in a
simple JSON-based vector store: each chunk's embedding vector, text, and
metadata were serialized into a JSON file, and similarity search was performed
by loading all vectors into memory and computing distances with NumPy. This
approach was sufficient for initial experiments, but its limitations became
clear as soon as the corpus began to grow:

\begin{itemize}
    \item \textbf{Scalability.} Loading the entire vector set into memory for
    every query is manageable for dozens of chunks, but becomes impractical as
    the corpus grows to hundreds or thousands.
    \item \textbf{No incremental updates.} Adding new documents required
    regenerating the entire JSON file and re-embedding all chunks, even if
    only one file had changed.
    \item \textbf{No metadata filtering.} The JSON store had no native support
    for filtering by document type, chapter, or year during retrieval; all
    such filtering had to be implemented ad hoc in application code.
    \item \textbf{No persistence guarantees.} The JSON file could be corrupted
    by concurrent writes or interrupted saves, with no transactional safety.
\end{itemize}

We therefore migrated to \textbf{ChromaDB}, an open-source vector database
designed for embedding storage and similarity search in AI applications.
ChromaDB provides persistent local storage (backed by SQLite for metadata and
an HNSW index for vectors), native metadata filtering, deterministic upsert
operations, and a Python API that integrates directly with custom embedding
functions~\cite{chromadb2024}. For a thesis prototype operating on a
domain-specific corpus of moderate size (approximately 2{,}000 chunks),
ChromaDB offered a good balance between simplicity and the features required
for a robust RAG pipeline. Managed cloud vector databases such as Pinecone or
Weaviate provide additional scalability features, but they also introduce
external dependencies and costs that were unnecessary at the scale of this
project.

The broader RAG literature consistently emphasizes the importance of
purpose-built vector storage for retrieval systems. Karpukhin~et~al.
demonstrated that dense passage retrieval with efficient vector indexing
(using FAISS in their case) dramatically outperforms sparse retrieval
methods such as BM25 for open-domain question answering~\cite{karpukhin2020dpr}.
The underlying architectural principle is the same whether the index is stored
in FAISS, ChromaDB, or another vector database: dense embeddings are
precomputed and then searched through approximate nearest-neighbor methods.
Our choice of ChromaDB reflects the practical need for a self-contained, locally
persistent store that supports metadata-filtered retrieval without requiring a
separate server process.

\subsubsection{Text Normalization}

Before chunking, each downloaded \texttt{.tex} file is normalized: line
endings are standardized (CRLF $\to$ LF), consecutive whitespace is
collapsed, and runs of three or more newlines are reduced to double newlines.
This helps ensure consistent chunk boundaries despite the formatting
differences introduced during the Gemini transcription step.

\subsubsection{Adaptive Chunking Strategy}

Not all documents in the corpus share the same internal structure. Correction
documents (Baccalaureate exams and exercise series) are usually organized
around discrete exercises, each with clearly numbered parts. Course material,
by contrast, contains longer theorem statements, proofs, and explanatory
passages whose meaning often extends across a larger span of text. A uniform
chunk size would therefore either split corrections in the middle of an
exercise or produce course chunks that are too small to preserve the relevant
context.

For this reason, the system uses \textit{adaptive chunking}: chunk size and
overlap are determined by document type. Correction-style documents
(\texttt{bac\_officiel}, \texttt{serie}) use a smaller chunk size of 1{,}500
characters with 200 characters of overlap, in order to preserve individual
exercise boundaries as much as possible. Course material (\texttt{cours})
uses a larger chunk size of 3{,}000 characters with 250 characters of
overlap, so that theorem statements and their proofs remain together whenever
possible. A default fallback of 1{,}800 characters with 200 characters of
overlap is used for documents that do not match either category.

The chunking algorithm itself prefers natural text boundaries over arbitrary
character positions. When a chunk reaches its target size, the algorithm
searches the last 40\% of the segment for the best available break point,
prioritizing, in order, paragraph breaks (double newline), line breaks
(single newline), sentence endings (period followed by space), and finally
any whitespace. This boundary-aware splitting reduces the likelihood of
cutting through a mathematical derivation or interrupting a logical argument
mid-sentence.

This approach is in line with recent findings by Zhao~et~al., who showed that
structure-aware chunking significantly outperforms naive fixed-size splitting
for RAG retrieval quality~\cite{zhao2024financial_chunking}. Chen~et~al.
further proposed Mix-of-Granularity, a framework that dynamically selects
optimal chunk granularity based on query characteristics, reinforcing the idea
that no single chunk size is optimal for heterogeneous document
collections~\cite{chen2024mog}.

\subsection{Embedding Model Selection}
\label{subsec:embedding-selection}

The quality of a dense retrieval system is ultimately constrained by the
quality of its embedding model. If two semantically different queries are
mapped to very similar vectors, the retrieval stage cannot distinguish between
them, and later steps in the pipeline have little chance of recovering that
lost distinction. This issue is especially important in mathematical domains,
where notational differences that look small such as $v(n)$ versus $u(n)$, or
$\sin(x)$ versus $\cos(x)$ can correspond to entirely different meanings.

\subsubsection{Initial Experience with Google Embedding Models}

Our first implementation used a Google-provided embedding model, accessed
through the Vertex AI Embeddings API. During testing, we noticed that this
model did not always distinguish reliably between queries involving closely
related mathematical sequences. For example, a question about the convergence
of a sequence $v(n)$ could retrieve chunks discussing $u(n)$, and vice versa,
even though these referred to different sequences with different recursive
definitions and convergence properties in a given exercise. In hindsight, this
is not especially surprising: general-purpose text embedding models are
trained mainly on natural-language data, where a one-character change often
does not alter the overall meaning of a passage. In mathematics, by contrast,
such differences may be decisive.

This observation is consistent with the broader literature on the limitations
of text embeddings for mathematical content. Research on mathematical
information retrieval has shown that mathematical expressions have
inherently two-dimensional spatial structures and may appear in several
semantically equivalent forms, which makes them fundamentally different from
linear natural-language text~\cite{mansouri2019tangent, zanibbi2016math_retrieval}.
Standard text embedding models, which are optimized for natural-language
similarity, tend to treat mathematical symbols as rare tokens with weakly
learned representations. The result can be poor discrimination between
expressions that are structurally similar but semantically distinct~\cite{pfahler2020math_embeddings}.

It should be noted that the specific observation about poor discrimination
between $v(n)$ and $u(n)$ comes from our own development experience rather
than from a controlled benchmark. We did not conduct a formal ablation study
comparing multiple embedding models on a mathematical retrieval dataset. The
decision to switch models was therefore based on recurring qualitative
retrieval failures observed during iterative system development.

\subsubsection{Migration to BGE-M3}

We replaced the Google embedding model with \textbf{BGE-M3}
(\textit{Multi-Linguality, Multi-Functionality, Multi-Granularity}), developed
by the Beijing Academy of Artificial Intelligence
(BAAI)~\cite{chen2024bge_m3}. BGE-M3 is built on an XLM-RoBERTa backbone,
continually pre-trained with RetroMAE, and supports input sequences of up to
8{,}192 tokens. It achieves state-of-the-art results on multilingual
(MIRACL), cross-lingual (MKQA), and long-document retrieval benchmarks.

Several characteristics of BGE-M3 make it particularly suitable for our use
case:

\begin{enumerate}
    \item \textbf{Multilingual support.} The Tunisian Baccalaureate
    mathematics corpus is written in French with embedded \LaTeX{} mathematical
    notation and occasional Tunisian Arabic (Derja) terms. BGE-M3 supports
    more than 100 languages, including French and Arabic, and was trained on
    multilingual retrieval tasks, which makes it well suited to the
    code-switching present in our documents.

    \item \textbf{Long context window.} With a maximum input length of 8{,}192
    tokens, BGE-M3 can embed our larger course-material chunks (up to 3{,}000
    characters) without truncation, preserving the full semantic content of
    theorem-length passages.

    \item \textbf{Multi-granularity training.} The model was trained on inputs
    ranging from short sentences to long documents, which fits the
    heterogeneous nature of our corpus, where short exercise statements appear
    alongside much longer course sections.

    \item \textbf{Dense retrieval quality.} BGE-M3 simultaneously supports
    dense, sparse, and multi-vector retrieval. In the present system, we use
    only the dense retrieval mode by extracting the \texttt{dense\_vecs}
    output. Even in this simpler setting, it provided better discrimination on
    our mathematical content than the previous model. In particular, the
    earlier $v(n)$/$u(n)$ confusion no longer appeared in the qualitative
    retrieval checks carried out during development.
\end{enumerate}

The embedding function is implemented as a custom ChromaDB
\texttt{EmbeddingFunction} class that wraps the \texttt{BGEM3FlagModel} from
the FlagEmbedding library. The model is loaded once as a singleton and reused
across embedding calls, with automatic GPU detection (\texttt{fp16=True}
when CUDA is available, \texttt{fp16=False} on CPU) for hardware-adaptive
performance.

\subsection{RAG Retrieval and Generation Engine}
\label{subsec:retrieval-engine}

The core runtime pipeline is implemented in the \texttt{rag\_engine.py} module,
which encapsulates the full retrieve--compile--generate cycle in a single
class (\texttt{TunisianMathRAG}). The module is deliberately
\textit{UI-agnostic}: it exposes a \texttt{.query(question, mode)} method
that returns a structured \texttt{QueryResult} dataclass .

\subsubsection{Two-Stage Retrieval}
\label{subsubsec:two-stage}

The retrieval strategy is one of the most distinctive design features of the
RAG engine. Rather than performing a single undifferentiated similarity search
over the full corpus, the engine uses a \textit{two-stage retrieval} pipeline
that gives priority to correction-style documents before falling back to
course material.

\textbf{First pass: correction exemplars.}
The first retrieval pass searches only among correction-type documents
(\texttt{type} $\in$ \{\texttt{bac\_officiel}, \texttt{serie}\}
with \texttt{is\_solution} = \texttt{true}), retrieving up
to $k_1 = 10$ candidates. The underlying idea is pedagogical: when a student
asks a question, the most useful reference is often a previously corrected
exercise that is similar in topic and structure. Such a correction provides
not only the mathematical content of the answer, but also a model of the
expected redaction style, including the sequence of steps, the use of logical
connectors (``On a\ldots'', ``Or\ldots'', ``Donc\ldots''), and the level of
detail typically expected in a Baccalaureate correction. By placing these
examples in the model context, the system encourages the LLM to reproduce the
method and style of a relevant correction while adapting it to the student's
question.

\textbf{Second pass: course-material fallback.}
If the best distance returned by the first pass exceeds a threshold , suggesting that no sufficiently similar correction is available, the
engine performs a second retrieval pass over course material
(\texttt{type} = \texttt{cours}), retrieving up to
$k_2 = 8$ candidates. These documents provide the theoretical support needed
to build a solution from definitions, properties, and theorem statements. In
this case, the prompt instructs the LLM to construct the answer from the
retrieved course material while preserving the expected Baccalaureate style.

If neither pass returns sufficiently relevant material, the engine performs an
unfiltered search over the corpus as a last resort, using a more relaxed
distance threshold.

\textbf{Pedagogical motivation.}
This two-stage design reflects a deliberate pedagogical distinction. In the
Tunisian Baccalaureate setting, there is a real difference between knowing the
theory and knowing how to write a solution in the expected exam format. A
correction exemplar shows the student how a solution is structured and
presented; a course reference provides the underlying mathematical tools. By
separating these two retrieval passes, the system can draw on the type of
material that is most useful for the query instead of mixing corrections and
theory indiscriminately.

\subsubsection{Similarity Thresholds}

The decision between retrieval stages is governed by two distance thresholds
applied to the L2 distances returned by ChromaDB's HNSW index
(lower distance = higher similarity):

\begin{description}
    \item[\texttt{SIMILARITY\_GOOD\_THRESHOLD} = 1.2]
    If the best distance from the first pass is at or below this value, the
    system treats the correction match as strong (Case~A) and uses only the
    correction-pass results, supplemented by companion exercise statements.

    \item[\texttt{SIMILARITY\_FALLBACK\_THRESHOLD} = 1.6]
    Documents with distances above this value are discarded entirely, regardless
    of which retrieval pass produced them. This helps prevent semantically
    irrelevant material from entering the model context and degrading answer
    quality through noise.
\end{description}

These thresholds were determined empirically during development. We inspected
the distribution of L2 distances returned by BGE-M3 embeddings across a range
of test queries and observed that distances below roughly 1.2 tended to
correspond to genuinely relevant corrections, whereas distances above roughly
1.6 more often returned material from unrelated topics. The intermediate range
(1.2--1.6) typically corresponded to partially relevant results that shared a
chapter or exercise type without matching the exact mathematical content of
the query.

We do not claim that these thresholds are theoretically optimal or directly
transferable to other embedding models or corpora. They are
implementation-specific parameters calibrated to the behavior of BGE-M3 on the
present Tunisian Baccalaureate corpus, and would likely need to be adjusted if
the embedding model, corpus size, or document domain were changed.

\subsubsection{Context Compilation}

The selected documents, whether they come from one retrieval pass or both, are
formatted into XML-tagged context blocks before being inserted into the LLM
prompt. Each block contains the document's GCS URI, metadata labels
(\texttt{type}, \texttt{chapter}, \texttt{year}), a role tag
(\texttt{[ÉNONCÉ]} for exercise statements, \texttt{[CORRECTION]} for
solutions), and the document content truncated to a maximum of 2{,}000
characters. The total context is capped at 14{,}000 characters so that it
remains within the model's effective context window while still leaving room
for the system prompt and the student's question.

\subsubsection{Prompt Construction and Generation}

The system prompt defines the persona of ``Monsieur Tounsi'', a 
Tunisian mathematics teacher, and encodes several constraints:

\begin{itemize}
    \item \textbf{Anti-hallucination rules:} The model must use \textit{only}
    information present in the retrieved context. If the context is
    insufficient, it must say so explicitly and provide only a method outline.
    \item \textbf{Mimicry decision tree:} In Case~A (correction found), the
    model is instructed to copy the method and redaction style of the retrieved
    correction. In Case~B (course material only), it must construct the
    solution from retrieved theorems.
    \item \textbf{Syllabus guard:} Methods outside the Tunisian
    Baccalaureate curriculum (L'Hôpital's rule, Taylor series, matrix
    diagonalization, etc.) are explicitly forbidden.
    \item \textbf{Derja sandwich:} The answer begins and ends with short
    motivational phrases in Tunisian Arabic dialect, while the mathematical
    body remains in formal academic French.
    \item \textbf{Anti-injection rules:} The model is warned that the context
    may contain adversarial text and must never follow instructions embedded
    inside retrieved documents.
\end{itemize}

The user prompt includes the interaction mode (correction for a dry,
exam-style answer or coaching  for a more pedagogical explanation),
the context blocks, the student's question, and final instructions requiring explicit source citation.


\subsubsection{Confidence Levels and Observability}

Every query returns a \texttt{QueryResult} dataclass that stores not only the
generated answer, but also detailed observability information: the list of
documents retrieved in each pass, their distances and metadata, the retrieval
case (A or B), timing breakdowns (retrieval time, generation time, total
time), and a confidence level.

The confidence level is derived from retrieval quality:
\begin{itemize}
    \item \textbf{Fort} (high): the best distance is $\leq 1.2$ and the
    total context exceeds 5{,}000 characters, indicating a strong match with
    substantial supporting material.
    \item \textbf{Moyen} (medium): the best distance is $\leq 1.6$ but
    either the match is weaker or the context is limited.
    \item \textbf{Faible} (low): no documents were retrieved, or all
    distances exceed the fallback threshold.
\end{itemize}

This observability layer was important for the thesis evaluation because it
made it possible to inspect retrieval decisions in detail, verify which
documents had been selected, and analyze failure cases more systematically.

\subsection{Paired Retrieval of Exercise and Correction}
\label{subsec:paired-retrieval}

During testing of the RAG engine, we noticed a structural problem in some of
the retrieved contexts: the system could retrieve a correction chunk without
the corresponding exercise statement, or retrieve an exercise statement
without its correction. This happens because chunking may separate the two,
and similarity search operates on individual chunks rather than on the
exercise--correction pair as a whole.

For mathematical reasoning, this is a real limitation. A correction is not
fully self-contained  it answers a specific exercise statement whose
conditions, notation, and variable names determine how the solution should be
read. Without that statement, neither the language model nor a human reader
can fully understand what the correction is addressing. The reverse problem is
also unhelpful: an exercise statement on its own gives the problem, but not
the worked method that the student is expected to learn from.

\subsubsection{Companion Fetch Mechanism}

To address this issue, the retrieval pipeline includes a \textit{companion
fetch} mechanism. After the two-stage retrieval has selected its final set of
correction documents, the engine performs a metadata-based lookup in order to
find the corresponding exercise statement for each selected correction:

\begin{enumerate}
    \item For each correction chunk in the selected results, the engine
    extracts the relevant metadata fields, in particular
    \texttt{chapter}, \texttt{type}, \texttt{year}, and \texttt{exo\_id}.

    \item Using these values, it builds a metadata filter that requests
    chunks with the same identifiers but with
    \texttt{is\_solution} = \texttt{false}, that is, exercise statements
    rather than corrections.

    \item This filter is executed through ChromaDB's
    \texttt{collection.get()} method, which performs a direct SQLite metadata
    lookup instead of a vector similarity search. In practice, this is both
    fast and more reliable for this purpose than re-running filtered
    similarity queries.

    \item The retrieved exercise chunks (up to 3 per correction group) are
    then prepended to the final context so that they appear before their
    corresponding corrections.
\end{enumerate}

During context compilation, these companion exercise chunks are labeled with
\texttt{[ÉNONCÉ]} and assigned a special distance marker
(\texttt{companion (énoncé)}) to distinguish them from chunks returned by
similarity search. This makes their role explicit: they are not included
because they independently matched the query, but because they complete the
context of a retrieved correction.

\subsubsection{Significance for Mathematical RAG}

This companion fetch mechanism does not appear to be explicitly described in
the existing RAG literature, which usually assumes that each retrieved chunk
is independently useful. In general-purpose question-answering systems
such as those described by Lewis et al.~\cite{lewis2020rag} and
Karpukhin et al.~\cite{karpukhin2020dpr}, that assumption is often
reasonable: a retrieved Wikipedia paragraph can usually be interpreted on its
own. In the present setting, however, the retrieved material comes from
structured pedagogical documents in which exercises and corrections form
logical pairs. In such a corpus, treating every chunk as fully independent is
not always appropriate.

The closest related idea in the literature is \textit{parent-document
retrieval}, where a system indexes smaller chunks for retrieval precision but
returns a larger parent document to restore context~\cite{gao2024rag_survey}.
The companion fetch mechanism used here can be understood as a specialized
version of that general pattern. Instead of returning a generic parent
document, the system retrieves the specific structural counterpart of the
selected chunk, namely the linked exercise statement or correction, identified
through metadata matching.

This design is therefore best described as an implementation-specific
adaptation for mathematical RAG in an educational setting. To our knowledge,
prior work has not explicitly implemented exercise--correction companion
retrieval in a tutoring system of this kind.
\subsection{Connection to the Tunisian Baccalaureate Context}
\label{subsec:bac-context}

The design decisions described in this section were shaped by the specific
requirements of generating Tunisian Baccalaureate-style mathematical
solutions. The system was not intended as a generic mathematics assistant, but
as a domain-adapted tutor for a particular exam context. Four characteristics
of that context were especially important:

\begin{enumerate}
    \item \textbf{Curriculum boundaries.} The Tunisian Baccalaureate
    (Section Mathématiques) covers a fixed set of chapters, and examiners
    penalize solutions that rely on methods outside the official program. The
    system therefore has to remain within those boundaries. In the
    implementation, this is supported both by the syllabus guard in the prompt
    and by metadata-based retrieval restricted to curriculum-aligned material.

    \item \textbf{Formal redaction conventions.} Baccalaureate corrections
    follow a recognizable rhetorical style: numbered steps, logical
    connectors in formal French (``On a\ldots'',
    ``D'après le théorème de\ldots'', ``Donc\ldots''), and explicit concluding
    statements. The two-stage retrieval strategy gives priority to correction
    exemplars for precisely this reason, so that the LLM can imitate this
    style and produce answers that resemble official corrections rather than
    generic mathematical explanations.

    \item \textbf{Exercise--correction dependency.} In Tunisian mathematical
    pedagogy, corrections are typically read together with their exercises.
    The companion fetch mechanism preserves this paired structure in the
    retrieved context, so that the LLM has access to the full logical unit
    needed to generate a coherent answer.

    \item \textbf{Bilingual and dialectal context.} The corpus is primarily
    in French with \LaTeX{} mathematical notation, but student questions may
    include Tunisian Arabic (Derja), and the intended pedagogical style also
    includes short motivational phrases in Derja. The choice of BGE-M3, with
    multilingual training covering both French and Arabic, supports this
    bilingual operating context.
\end{enumerate}


\section{Prompt-Only Baseline}
\subsubsection{Role and Rationale}

The prompt-only baseline serves as the non-retrieval comparison system in the
experimental design. It uses the same Gemini model as the RAG system, with the
same temperature ($T = 0.15$) and the same maximum output length
(4{,}096 tokens), but it never queries the vector database.

Instead, all task guidance is provided directly in the prompt. This includes
the expected answer style, the curriculum boundaries, and the pedagogical
framing of the response. In other words, the system must rely entirely on the
model's parametric knowledge together with the instructions encoded in the
prompt itself.

Including such a baseline is important because modern language models can
often produce reasonable mathematical answers even without retrieval, provided
that the prompt is well structured~\cite{brown2020gpt3, wei2022chain_of_thought}.
Without this comparison point, it would be difficult to determine whether the
performance of the RAG system comes from the retrieved corpus or simply from
the underlying generative model. The prompt-only baseline therefore makes it
possible to measure the added value of retrieval more clearly.

For that reason, the baseline was kept as close as possible to the RAG system
in all other respects. Both systems use the same ``Monsieur Tounsi'' persona,
the same interaction modes (\textit{correction} and \textit{coaching}), and
the same general output constraints. Both also expose the same public API
through a \texttt{query(question, mode)} method. The main difference is thus
deliberately limited to the source of domain knowledge: retrieved external
documents in the RAG system, versus prompt-encoded guidance in the prompt-only
baseline.

\subsubsection{Prompt Architecture}

The system prompt is organized into ten functional components that are
concatenated into a single prompt string at inference time. This modular
organization follows the general idea of \textit{prompt
patterns}~\cite{white2023prompt_patterns}, in the sense that different parts
of the prompt are used to handle different constraints of the task.

Some components define the pedagogical and stylistic identity of the system,
while others are there to reduce recurrent failure modes observed during
development. Taken together, they form the practical backbone of the
prompt-only baseline.

\begin{description}
    \item[Component~1: System persona.] Defines ``Monsieur Tounsi'' as a
    veteran Tunisian mathematics teacher with more than twenty years of
    Baccalaureate preparation experience. This persona framing was chosen to
    steer the model toward a more domain-appropriate register and vocabulary,
    in line with the idea that role-based prompts can improve task
    adherence~\cite{white2023prompt_patterns}.

    \item[Component~2: Curriculum inventory.] Enumerates all ten chapters of
    the Section Mathématiques program, from suites numériques to logarithme
    and exponentielle. This is the main knowledge-orientation component.
    Because the system does not retrieve external documents, the prompt must
    specify which part of mathematics the model is expected to stay within.
    The inventory itself was compiled from the official Tunisian
    Baccalaureate program and checked against the source material used in the
    corpus. It is therefore an implementation-specific artifact rather than a
    published benchmark resource.

    \item[Component~3: Reasoning guardrails.] Defines a six-step internal
    checklist (identify, recall, verify, plan, solve, check) that the model
    is instructed to follow before producing the final answer. This component
    is inspired by the broader chain-of-thought prompting
    literature~\cite{wei2022chain_of_thought}, but is implemented here as an
    explicit procedural scaffold rather than through few-shot examples.

    \item[Component~4: Output structure.] Specifies the six-part answer
    format (identification, rappel du cours, méthode, application,
    conclusion, vérification) together with the expected \LaTeX{} formatting
    conventions. Both interaction modes, \textit{correction} and
    \textit{coaching}, follow this overall structure, although coaching mode
    allows more explanatory phrasing.

    \item[Component~5: Hallucination control.] Because the prompt-only system
    has no retrieved context to anchor its answers, this component is stricter
    than in the RAG version. It instructs the model not to invent specific
    exercise numbers, textbook references, or exam sources unless it is
    certain they exist, and to prefer more general wording when uncertainty is
    high. This was introduced as an implementation-specific response to early
    testing, where the model occasionally produced plausible-sounding but
    nonexistent references.

    \item[Component~6: Syllabus guard.] Lists methods that are explicitly
    outside the Baccalaureate program, such as L'Hôpital's rule, Taylor
    series, matrix diagonalization, Laplace or Fourier transforms, and
    improper integrals. The model is instructed to refuse these politely and
    redirect the answer toward a Bac-compatible alternative.

    \item[Component~7: Derja sandwich.] Specifies that each answer should open
    and close with a short motivational phrase in Tunisian Arabic (Derja),
    while the mathematical body remains in formal academic French. This
    convention was not taken from the literature; it reflects a pedagogical
    style familiar from real teacher--student interaction in the Tunisian
    context.

    \item[Component~8: Self-verification block.] Requires the model to add a
    visible verification step at the end of the answer, for example by
    substituting a computed value back into an equation. The goal is both to
    encourage internal checking and to make verification part of the student's
    visible solving routine.

    \item[Component~9: Refusal policy.] Defines how the system should respond
    to requests that fall outside the intended scope, including questions from
    another educational level, requests to generate exam papers, or prompts
    that ask explicitly for out-of-program methods.

    \item[Component~10: Confidence calibration.] Maps the model's internal
    sense of certainty to three labels (fort, moyen, faible), which are shown
    at the end of the answer. Unlike the RAG system, this calibration cannot
    rely on retrieval distances and is therefore based only on question
    characteristics and prompt-internal heuristics.
\end{description}

This ten-component structure is an implementation-specific design choice. We
are not aware of a published reference that decomposes mathematical tutoring
prompts in exactly this way. The structure emerged gradually during
development, as repeated failure modes such as curriculum violations,
hallucinated references, and inconsistent formatting made it necessary to add
more targeted prompt components.

\subsubsection{Confidence Estimation Without Retrieval}

A clear limitation of the prompt-only baseline is that it has no retrieval
signal on which to base a confidence estimate. In the RAG system, confidence
can be derived from the distance of the best-matching document together with
the amount of retrieved context (Section~\ref{subsec:retrieval-engine}). The
prompt-only system has no equivalent source of evidence.

As a practical workaround, we implemented a simple keyword-based heuristic.
If the question contains terms strongly associated with standard Bac exercise
types, such as ``récurrence'', ``limite'', ``intégrale'', ``PGCD'', or
``Bézout'', the system reports high confidence on the assumption that the
model's parametric knowledge of these common topics is more likely to be
sufficient. Very short or ambiguous questions are assigned low confidence.
All remaining cases default to medium.

This heuristic is intentionally simple. It is presented as an
implementation-specific approximation rather than as a principled confidence
calibration method. Its main purpose is to provide a confidence signal that
can be recorded in the same evaluation framework as the other two systems.

\subsubsection{Limitations as a Baseline}

The prompt-only system makes a fairly strong assumption: that the language
model's parametric knowledge, when guided by a carefully structured prompt, is
sufficient to produce Bac-style mathematical answers.

In practice, we expected this approach to work better for standard and
frequently occurring exercise types, which are more likely to resemble
patterns seen during pretraining, than for less common topics or questions
that depend heavily on the exact notation, wording, and correction conventions
of the Tunisian program. The evaluation
(Section~\ref{sec:experiments}) tests to what extent this expectation is borne
out in practice.

\section{Hybrid Engine}

\subsubsection{Motivation}

The RAG system and the prompt-only baseline correspond to two opposite design
choices. The RAG system always retrieves and always conditions its answer on
retrieved context, even when the retrieved material is only weakly relevant.
The prompt-only system does the reverse: it never retrieves and relies
entirely on the model's parametric knowledge, even when highly relevant
corrections exist in the corpus. In practice, neither extreme is ideal for
all queries.

This intuition is reflected in recent work on adaptive retrieval. Several
studies argue that the question is not only how retrieval should be performed,
but also when it should be trusted. Yan~et~al.~\cite{yan2024practical_rag},
for example, propose Corrective RAG (CRAG), in which retrieval quality is
evaluated before deciding whether corrective actions are needed. Asai
et~al.~\cite{asai2023self_rag} introduce Self-RAG, where the model itself can
decide whether retrieval is necessary and can assess the usefulness of the
retrieved evidence. Although these systems are more sophisticated than the one
implemented here, they share the same basic insight: a fixed retrieval policy
is often suboptimal.

The hybrid engine developed in this thesis adopts a simpler version of this
idea. It does not train a retrieval critic and does not add an external web
search fallback. Instead, it routes each query to one of three generation
strategies based on the L2 distance of the best retrieved document, a signal
that is already produced by the retrieval pipeline and can therefore be used
at no additional cost.

\subsubsection{Adaptive Routing Logic}

The hybrid engine is not implemented as a separate retrieval system. It is a
thin routing layer built on top of the existing RAG engine. Concretely, it
calls the same two-stage retrieval procedure
(Section~\ref{subsec:retrieval-engine}), receives the same candidate
documents together with their distances, and then decides how the final answer
should be generated. The routing function evaluates the distance of the
best-matching \textit{real} retrieval result, while excluding companion
documents, since those are inserted for context completeness and carry a
synthetic distance marker rather than a genuine similarity score.

Three cases are distinguished:

\begin{description}
    \item[Case~A: Strong retrieval] (best distance $\leq 1.2$).
    The query closely matches one or more corrections in the corpus. In this
    situation, the system behaves like the standalone RAG engine described in
    Section~\ref{subsec:retrieval-engine}: the prompt instructs the model to
    rely only on the retrieved context, to imitate the style of the retrieved
    corrections, and to cite sources explicitly. The same anti-hallucination
    and anti-injection constraints remain in place.

    \item[Case~B: Partial retrieval] ($1.2 <$ best distance $\leq 1.6$).
    The system has retrieved documents that are not irrelevant, but are also
    not strong enough to justify a purely retrieval-grounded answer. In this
    case, the retrieved material is kept in the context, but the prompt is
    also supplemented with the curriculum inventory and syllabus guard from
    the prompt-only system. The model is therefore allowed to combine partial
    retrieval evidence with parametric knowledge, while being instructed to
    make source attribution explicit where possible. This mixed-source
    behavior is an implementation-specific design choice intended to preserve
    useful retrieval without forcing the model to depend on it too strongly.

    \item[Case~C: No useful retrieval] (best distance $> 1.6$, or no
    documents returned).
    The retrieved material is considered too weak to be useful. In this case,
    the system discards the retrieved documents and falls back to the same
    strategy used by the prompt-only baseline.
\end{description}

This three-case routing scheme can be seen as a simplified practical analogue
of the retrieval-quality assessment discussed in CRAG~\cite{yan2024practical_rag}.
The difference is that our system uses fixed thresholds rather than a learned
quality estimator. That makes the implementation much simpler, but also means
that the thresholds have to be set empirically.

\subsubsection{Threshold Determination}

The two thresholds that define the boundaries between Cases~A, B, and C are
the same thresholds already used in the standalone RAG engine: 1.2 for the
Case~A/B boundary and 1.6 for the Case~B/C boundary
(Section~\ref{subsec:retrieval-engine}). They were selected empirically by
inspecting the distribution of L2 distances returned by BGE-M3 over a range of
test queries during development. Distances below 1.2 were usually associated
with genuinely relevant corrections, while distances above 1.6 tended to
return material from unrelated topics. The intermediate range generally
corresponded to partial matches.

As in the RAG system, no claim is made that these thresholds are theoretically
optimal or that they would transfer unchanged to another corpus or embedding
model. They should be understood as implementation-specific parameters tuned
to the behavior of BGE-M3 on the present Tunisian Baccalaureate corpus.

\subsubsection{Confidence Management}

The confidence signal in the hybrid engine depends on the routing case.

In \textbf{Case~A}, confidence is computed in the same way as in the
standalone RAG system, using retrieval quality and context volume. In
\textbf{Case~C}, confidence falls back to the same keyword-based heuristic as
in the prompt-only baseline.

\textbf{Case~B} required a separate decision. In this case, the answer is
partly supported by retrieved documents and partly supplemented by parametric
knowledge. For that reason, confidence is explicitly capped at
``moyen'' (medium), regardless of other signals. This cap is an
implementation-specific choice. It reflects the view that a mixed-source answer
should not be presented as highly reliable if the retrieval quality was not
strong enough to justify Case~A in the first place. In the configuration, this
appears as the parameter
\texttt{HYBRID\_CASE\_B\_MAX\_CONFIDENCE = "moyen"}.

\subsubsection{Relationship to Adaptive Retrieval Literature}

Within the broader literature on adaptive retrieval, the hybrid engine used
here occupies a fairly simple middle position. Compared with
CRAG~\cite{yan2024practical_rag}, which can trigger additional corrective
steps such as web search, our system remains closed: it can only use the
corpus and the model's own parametric knowledge. Compared with
Self-RAG~\cite{asai2023self_rag}, which relies on model-level retrieval and
critique tokens, our routing mechanism is much simpler and does not require
any model modification or additional training.

This simplicity is deliberate. For a bachelor thesis built around a fixed,
domain-specific corpus, the goal was not to propose a new adaptive retrieval
framework, but to test whether a lightweight quality-aware routing policy
could reduce two practical failure modes: forcing generation from weak
retrieval results, and ignoring relevant retrieved material when it is
available. Whether this improves performance in practice is examined in the
evaluation section (Section~\ref{sec:experiments}).
